### 2019-04-09 paper/main.tex
initial tex document for paper

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

\title{Optimizing biodiversity sampling by citizen scientists}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b,1]{Corey T. Callaghan}
\author[c,]{Alistair G. B. Poore} 
\author[b,a]{Richard E. Major}
\author[b,a]{Jodi J. L. Rowley}
\author[a,c]{William K. Cornwell}

\affil[a]{Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}
\affil[b]{Australian Museum Research Institute, Australian Museum, Sydney, 2000, NSW, Australia}
\affil[c]{Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}

% Please give the surname of the lead author for the running footer
\leadauthor{Callaghan}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Citizen scientists are increasingly monitoring biodiversity, but we need to better understand how to optimally sample biodiversity in time and space for intended outcomes, relying on citizen science data.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{The authors have no conflict of interest to declare.}

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: c.callaghan@unsw.edu.au}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{biodiversity $|$ conservation $|$ citizen science $|$ optimal sampling $|$ eBird}

\begin{abstract}
Citizen science is important. yata yata yata
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}

This work is really important \cite{theobald2015global}.

\section*{Results}
\section*{Discussion}

\matmethods{Please describe your materials and methods here. This can be more than one paragraph, and may contain subsections and equations as required. Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.

\subsection*{eBird data}
We used eBird data....

\subsection*{Parameter calculation}
For 2018, we
}

\showmatmethods{} % Display the Materials and Methods section

\acknow{We thank the countless citizen scientists who are contributing data that is continuously increasing our collective knowledge of biodiversity.}

\showacknow{} % Display the acknowledgments section

% Bibliography
\bibliography{refs}

\end{document}

### 2019-04-09 R/build_df_for_parameterizing_model.R

@@ -1,232 +1,0 @@
-### Create a dataframe for which modelling against the formula
-
-
-# packages
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(sf)
-library(data.table)
-library(geosphere)
-library(BBmisc)
-library(RANN)
-
-
-# load data
-load("Data/Greater_Sydney_data.RData")
-
-load("Data/Sampling data/NSW_birds.RData")
-
-load("Data/Spatial data/sydney_grids.RData")
-
-
-
-# assign each locality (and thereby sampling event identifier) 
-# a grid cell throughout Sydney
-sydney_grids <- sydney_grids %>%
-  rename(grid_id = id)
-
-# get list of unique sites
-sites <- GS_observations %>%
-  dplyr::select(LOCALITY_ID, LATITUDE, LONGITUDE) %>%
-  distinct()
-
-
-# convert points from FrogID into sf points
-# so that they can be used with the polygons
-crs <- "+proj=longlat +ellps=GRS80 +no_defs"
-
-sites_sf <- st_as_sf(sites, coords = c("LONGITUDE", "LATITUDE"), 
-                      crs = crs, agr = "constant")
-
-# use intersect to return a list of points and the associated
-# polygon they belong to - uses integer ids for rows and columns
-# note that it looks like 29 points aren't assigned to a LGA
-# not worried about this for now, but something to come back to in the future
-assigned_points <- as.data.frame(st_intersects(sites_sf, sydney_grids))
-
-# join the original dataframe with the intersected df
-# then remove the row and col ids
-sites_and_grids <- sites %>%
-  mutate(row.id = 1:nrow(.)) %>%
-  left_join(assigned_points, by="row.id") %>%
-  rename(grid_id = col.id) %>%
-  dplyr::select(-row.id)
-
-
-test <- sydney_grids %>%
-  left_join(., sites_and_grids, by="grid_id") %>%
-  left_join(., GS_observations, by="LOCALITY_ID") %>%
-  group_by(grid_id) %>%
-  summarise(N=length(unique(SAMPLING_EVENT_IDENTIFIER))) 
-
-
-
-
-
-date_specific_summary <- function(forecast_date) {
-
-# do it for a single date first
[…]

### 2019-04-09 R/build_df_for_parameterizing_model_for_grids.R

@@ -1,331 +1,0 @@
-### Create a dataframe for which modelling against the formula can be done
-### This means that for everyday throughout 2018 I will calculate what grids have
-### been sampled and the various aspects that might be useful in
-### calculating where to go sample next
-### this dataframe will then be put together with the leverage of every checklist
-### to see how well people are sampling
-### will do this for every day in 2018
-### so that the function is passed a date and then lapply over a list of dates
-
-
-# packages
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(sf)
-library(data.table)
-library(geosphere)
-library(BBmisc)
-library(RANN)
-library(nngeo)
-
-
-# load data
-load("Data/Greater_Sydney_data.RData")
-load("Data/Sampling data/NSW_birds.RData")
-load("Data/Spatial data/sydney_grids_1_km.RData")
-load("Data/Spatial data/sydney_grids_2_km.RData")
-load("Data/Spatial data/sydney_grids_5_km.RData")
-load("Data/Spatial data/sydney_grids_10_km.RData")
-load("Data/Spatial data/sydney_grids_25_km.RData")
-load("Data/Spatial data/sydney_grids_50_km.RData")
-
-
-
-# assign each locality (and thereby sampling event identifier) 
-# a grid cell throughout Sydney
-# will be necessary for merging back later
-# because the calculations will be on a grid scale
-# get list of unique sites (LOCALITY_ID)
-sites <- GS_observations %>%
-  dplyr::select(LOCALITY_ID, LATITUDE, LONGITUDE) %>%
-  distinct()
-
-
-# convert points into sf points
-# so that they can be used with the polygons
-crs <- 4326
-
-sites_sf <- st_as_sf(sites, coords = c("LONGITUDE", "LATITUDE"), 
-                      crs = crs, agr = "constant")
-
-
-date_specific_summary <- function(forecast_date, dataset, grid_size) {
-
-  
-# use intersect to return a list of points and the associated
-# polygon they belong to - uses integer ids for rows and columns
-assigned_points <- as.data.frame(st_intersects(sites_sf, dataset))
-
-# join the original dataframe with the intersected df
[…]

### 2019-04-09 R/create_model_df.R

@@ -1,182 +1,0 @@
-# An R script which reads in each of the day's sampling
-# summary, for each of the different grid cell sizes
-# and makes dataframes.
-# then combines those dataframes into the 'parameters' df
-
-library(dplyr)
-library(tidyr)
-
-
-
-# get a dataframe for modelling 
-# for each of the different grid cell sizes
-
-# 50 km first
-setwd("Data/Sampling dates 2018/50 km")
-
-file_list <- list.files()
-
-
-for (file in file_list){
-  
-  # if the merged dataset doesn't exist, create it
-  if (!exists("dataset")){
-    dataset <- readRDS(file)
-  }
-  
-  # if the merged dataset does exist, append to it
-  if (exists("dataset")){
-    temp_dataset <-readRDS(file)
-    dataset<-rbind(dataset, temp_dataset)
-    rm(temp_dataset)
-  }
-  
-}
-
-dataset1 <- dataset %>%
-  distinct() %>%
-  mutate(grid_size=50)
-
-rm(dataset)
-
-setwd('..')
-setwd('..')
-setwd('..')
-
-# 25 km
-setwd("Data/Sampling dates 2018/25 km")
-
-file_list <- list.files()
-
-
-for (file in file_list){
-  
-  # if the merged dataset doesn't exist, create it
-  if (!exists("dataset")){
-    dataset <- readRDS(file)
-  }
-  
-  # if the merged dataset does exist, append to it
-  if (exists("dataset")){
-    temp_dataset <-readRDS(file)
-    dataset<-rbind(dataset, temp_dataset)
-    rm(temp_dataset)
-  }
-  
-}
-
-dataset2 <- dataset %>%
-  distinct() %>%
-  mutate(grid_size=25)
-
-rm(dataset)
-
-setwd('..')
-setwd('..')
-setwd('..')
-
-# 10 km
-setwd("Data/Sampling dates 2018/10 km")
-
-file_list <- list.files()
-
-
-for (file in file_list){
-  
-  # if the merged dataset doesn't exist, create it
-  if (!exists("dataset")){
-    dataset <- readRDS(file)
-  }
-  
-  # if the merged dataset does exist, append to it
-  if (exists("dataset")){
-    temp_dataset <-readRDS(file)
-    dataset<-rbind(dataset, temp_dataset)
-    rm(temp_dataset)
-  }
-  
-}
-
-dataset3 <- dataset %>%
[…]

### 2019-04-09 R/create_raster_of_sydney.R

@@ -1,78 +1,0 @@
-# create a raster of the study area
-
-library(sf)
-library(dplyr)
-
-
-# load shapefile
-sydney <- st_read("Data/Spatial data/greater_sydney_shape/sydney.shp") %>%
-  st_transform(32756)
-
-
-# function to make a grid over Sydney region
-# grid size is defined in meters
-get_gridded_region_function <- function(grid_size) {
-  
-  sydney_grids <- sydney %>%
-    st_make_grid(., cellsize=c(grid_size, grid_size),
-                 crs=32756, what='polygons') %>%
-    st_intersection(sydney) %>%
-    st_cast("MULTIPOLYGON") %>%
-    st_sf() %>%
-    mutate(id = row_number())
-  
-  
-  centroids <- st_centroid(sydney_grids)
-  centroids <- st_transform(centroids, 4326)
-  sydney_grids <- st_transform(sydney_grids, 4326)
-  
-  coords <- do.call(rbind, st_geometry(centroids)) %>%
-    as_tibble() %>%
-    setNames(c("lon", "lat"))
-  
-  sydney_grids <- bind_cols(sydney_grids, coords) %>%
-    rename(grid_id = id)
-  
-  return(sydney_grids)
-  
-}
-
-# 50 km grids
-sydney_grids_50_km <- get_gridded_region_function(grid_size=50000)
-
-save(sydney_grids_50_km, file = "Data/Spatial data/sydney_grids_50_km.RData")
-
-
-# 25 km grids
-sydney_grids_25_km <- get_gridded_region_function(grid_size=25000)
-
-save(sydney_grids_25_km, file = "Data/Spatial data/sydney_grids_25_km.RData")
-
-
-# 10 km grids
-sydney_grids_10_km <- get_gridded_region_function(grid_size=10000)
-
-save(sydney_grids_10_km, file = "Data/Spatial data/sydney_grids_10_km.RData")
-
-
-# 5 km grids
-sydney_grids_5_km <- get_gridded_region_function(grid_size=5000)
-
-save(sydney_grids_5_km, file = "Data/Spatial data/sydney_grids_5_km.RData")
-
-
-# 2 km grids
-sydney_grids_2_km <- get_gridded_region_function(grid_size=2000)
-
-save(sydney_grids_2_km, file = "Data/Spatial data/sydney_grids_2_km.RData")
-
-
-# 1 km grids
[…]

### 2019-04-09 R/data_frame_testing.R

@@ -1,12 +1,0 @@
-library(tidyverse)
-
-lev<-readRDS("Data/Modelling data/leverage_results.RDS")
-param<-readRDS("Data/Modelling data/parameter_dataset.RDS")
-
-
-all<-left_join(lev,param)
-
-#should be a lot of 4's 
-sort(table(all$SAMPLING_EVENT_IDENTIFIER),decreasing = T)
-
-

### 2019-04-09 R/get_all_observations_from_greater_sydney.R

@@ -1,77 +1,0 @@
-# An R script to get all eBird data subsetted to the greater sydney region
-# also filtered to get rid of 'bad' checklists
-# it takes a long time for these filtering of the MariaDB to be done
-# so go get a coffee or something (or a beer!!)
-
-library(sf)
-library(dbplyr)
-library(dplyr, quietly = T, warn.conflicts = F)
-library(RMariaDB)
-get_con  <- function () {
-  dbConnect(RMariaDB::MariaDB(), host = 'KESTREL', dbname='ebird',user = Sys.getenv('userid'), password = Sys.getenv('pwd'))
-} 
-
-sites <- tbl(con, "sites")
-samples <- tbl(con, "samples")
-species <- tbl(con, "species")
-checklists <- tbl(con, "checklists")
-
-NSW_species <- sites %>% 
-  filter(STATE_PROVINCE == "New South Wales") %>% 
-  left_join(., checklists, by="LOCALITY_ID") %>% 
-  inner_join(., samples, by="SAMPLING_EVENT_IDENTIFIER") %>% 
-  collect(n=Inf)
-
-NSW_sites <- sites %>%
-  filter(STATE_PROVINCE == "New South Wales") %>%
-  collect(n=Inf)
-
-species_all <- species %>%
-  select(TAXONOMIC_ORDER, CATEGORY, COMMON_NAME, SCIENTIFIC_NAME) %>%
-  collect(n=Inf)
-
-## assign each observation to a Greater Capital City Statistical Areas (GCCSA)
-## This step requires download of shapefile from
-## here http://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/1270.0.55.001July%202016?OpenDocument )
-## we then read in the shapefile and assign each point to a particular GCCSA
-## this will be used for later visualizations and analyses
-
-
-# read in the shapefile
-aus_gccsa <- st_read("Data/Spatial data/gccsa_2016_aust_shape/GCCSA_2016_AUST.shp")
-
-# turn observations into sf object
-# only for unique lists
-points_sf <- st_as_sf(NSW_sites, coords = c("LONGITUDE", "LATITUDE"), crs=st_crs(aus_gccsa)) 
-
-# create metadata df
-# used to link the dfs back together later
-gccsa_metadata <- data.frame(gccsa_name=aus_gccsa[[2]], col.id=1:34)
-
[…]

### 2019-04-09 R/global_functions.R

@@ -1,25 +1,0 @@
-# global functions
-
-factory <- function(fun)
-  function(...) {
-    warn <- err <- NULL
-    res <- withCallingHandlers(
-      tryCatch(fun(...), error=function(e) {
-        err <<- conditionMessage(e)
-        NULL
-      }), warning=function(w) {
-        warn <<- append(warn, conditionMessage(w))
-        invokeRestart("muffleWarning")
-      })
-    list(res, warn=warn, err=err)
-  }
-
-
-
-.has <- function(x, what)
-  !sapply(lapply(x, "[[", what), is.null)
-hasWarning <- function(x) .has(x, "warn")
-hasError <- function(x) .has(x, "err")
-isClean <- function(x) !(hasError(x) | hasWarning(x))
-value <- function(x) sapply(x, "[[", 1)
-cleanv <- function(x) sapply(x[isClean(x)], "[[", 1)

### 2019-04-09 R/initial_script_for_formula.R

@@ -1,458 +1,0 @@
-# An initial R script, attempting to develop a 'formula'
-# which can provide value to a given site
-# relative to other sites
-
-
-
-# packages
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(sf)
-library(data.table)
-library(geosphere)
-library(BBmisc)
-library(RANN)
-
-# read data in
-load("Data/Sampling data/NSW_birds.RData")
-
-
-# select ten 'locations/sites'
-# this is done by just browsing on th eBird 
-# website for now
-# eventually
-# if we get a formula that 'works'
-# then we will want to come back to how to define a site
-example_sites <- c("L945869", "L2557723", "L8083126", "L1030678", "L915566", 
-                   "L1300136", "L5076722", "L2444301", "L3007885", "L5146094")
-
-
-# get data out for those ten sites
-example_data <- NSW_ebird %>%
-  dplyr::filter(PROTOCOL_TYPE != "Incidental") %>%
-  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
-  dplyr::filter(OBSERVATION_DATE >= "2010-01-01") %>%
-  dplyr::filter(LOCALITY_ID %in% example_sites)
-
-
-# now need to calculate the following variables for
-# each site
-# 1.) Whether a site was sampled or not
-# 2.) distance from the nearest sampled site (nearest neighbor)
-# 3.) The median waiting time between checklists at the site
-# 4.) days between first and last observation
-# 5.) The median waiting time at the nearest neighbor
-# 6.) Days since last sampling event
-
-
-# 1.) Whether a site was sampled or not
-# all sites
-all_sites <- NSW_ebird %>%
-  dplyr::filter(LOCALITY_ID %in% example_sites) %>%
-  dplyr::select(LOCALITY_ID) %>%
-  distinct() 
-
-# sampled sites
-sampled_sites <- example_data %>%
-  dplyr::select(LOCALITY_ID) %>%
-  distinct() %>%
-  mutate(sampled = "yes")
-
-sampled_or_not <- all_sites %>%
-  left_join(., sampled_sites, by="LOCALITY_ID") %>%
-  replace_na(list(sampled="no"))
-
-
-# 2.) distance from the nearest sampled site
-# not sure how well this approach will scale for a large
-# number of 'sites'
-# but it works for now
[…]

### 2019-04-09 R/modelling.R

@@ -1,317 +1,0 @@
-### Modelling script
-
-# packages
-library(dplyr)
-library(ggplot2)
-library(BBmisc)
-library(ggcorrplot)
-library(corrplot)
-library(lme4)
-library(standardize)
-library(tibble)
-library(visreg)
-library(arm)
-
-# load data
-lev <- readRDS("Data/Modelling data/leverage_results.RDS")
-
-params <- readRDS("Data/Modelling data/parameter_dataset.RDS")
-
-analysis_data <- params %>%
-  left_join(., lev) %>%
-  mutate(
-    sampled_numeric=case_when(
-    sampled=="yes" ~ 1,
-    sampled=="no" ~ 0)
-    )
-
-# make df for modelling
-hist(analysis_data$DATE_dfbetas, breaks=30)
-
-hist(log(analysis_data$DATE_dfbetas), breaks=30)
-
-# rescale variables
-analysis_data <- analysis_data %>%
-  group_by(grid_size) %>%
-  mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
-  mutate(s.duration=arm::rescale(duration)) %>%
-  mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
-  mutate(s.days_since=rescale(days_since)) %>%
-  mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
-  mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time))
-
-# look at correlation among predictor variables
-cor_object5 <- analysis_data %>%
-  dplyr::filter(grid_size==5) %>%
-  dplyr::select(s.median_waiting_time, s.duration, s.Number_of_days_sampled, 
-                s.days_since, s.dist_km_nn, s.neighbor_waiting_time) %>%
-  ungroup()
-
-ggcorrplot(cor(na.omit(cor_object5[2:7])))+
-  ggtitle("5 km grid size")
-
-# look at correlation among predictor variables
-cor_object10 <- analysis_data %>%
-  dplyr::filter(grid_size==10) %>%
-  dplyr::select(s.median_waiting_time, s.duration, s.Number_of_days_sampled, 
-                s.days_since, s.dist_km_nn, s.neighbor_waiting_time) %>%
-  ungroup()
-
-ggcorrplot(cor(na.omit(cor_object10[2:7])))+
-  ggtitle("10 km grid size")
-
-
-# look at correlation among predictor variables
[…]

### 2019-04-09 R/plot_leverage.R

@@ -1,122 +1,0 @@
-## An R script to plot predicted leverage
-## Will do it for 5 km grid cell size first
-
-library(dplyr)
-library(sf)
-library(ggplot2)
-library(wesanderson)
-
-# 5 km first
-setwd("Data/predicted_leverage/5_km")
-
-file_list <- list.files()
-
-
-for (file in file_list){
-  
-  # if the merged dataset doesn't exist, create it
-  if (!exists("dataset")){
-    dataset <- readRDS(file)
-  }
-  
-  # if the merged dataset does exist, append to it
-  if (exists("dataset")){
-    temp_dataset <-readRDS(file)
-    dataset<-rbind(dataset, temp_dataset)
-    rm(temp_dataset)
-  }
-  
-}
-
-setwd('..')
-setwd('..')
-setwd('..')
-
-load("Data/Spatial data/sydney_grids_5_km.RData")
-
-
-leverage_5 <- dataset %>%
-  distinct() %>%
-  mutate(Value=exp(.fitted)) %>%
-  group_by(grid_id, dynamic_date) %>%
-  sample_n(1) %>%
-  ungroup() 
-
-
-
-df <- leverage_5 %>%
-  dplyr::filter(dynamic_date==date)
-
-map_df <- sydney_grids_5_km %>%
-  right_join(., df, by="grid_id")
-
-plasma_pal <- c("red", viridis::plasma(n = 6))
-
-ggplot()+
-  geom_sf(data=map_df, aes(fill=Value))+
-  theme_bw()+
-  theme(panel.grid.major=element_line(color="transparent"))+
-  theme(axis.text=element_blank())+
-  theme(axis.ticks=element_blank())+
-  scale_fill_gradientn(colours=pal, name="Value", trans="log")
-  
-# potential figure for paper
-example_dates <- c("2018-01-01", "2018-03-01", "2018-05-01",
-                   "2018-07-01", "2018-09-01", "2018-11-01")
-
-df <- leverage_5 %>%
-  dplyr::filter(as.character(as.Date(dynamic_date)) %in% example_dates)
-
-map_df <- sydney_grids_5_km %>%
-  right_join(., df, by="grid_id") %>%
-  st_simplify()
-
-
-ggplot()+
-  geom_sf(data=map_df, aes(fill=Value))+
-  theme_bw()+
-  theme(panel.grid.major=element_line(color="transparent"))+
-  theme(axis.text=element_blank())+
[…]

### 2019-04-09 R/population_trend_model.R

@@ -1,262 +1,0 @@
-## A function to apply for every species to assess 'population trends'
-
-# packages
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(ggplot2)
-library(mgcv)
-library(lme4)
-library(car)
-library(readr)
-
-
-# load data
-load("Data/Greater_Sydney_data.RData")
-
-clements <- read_csv("Data/Clements-Checklist-v2018-August-2018.csv") %>%
-  rename(COMMON_NAME = `English name`)
-
-# filter the dataset
-# remove any incomplete checklists
-GS_observations <- GS_observations %>%
-  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
-  dplyr::filter(CATEGORY %in% c("species", "issf", "domestic")) %>%
-  dplyr::filter(COMMON_NAME != "Indian Peafowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Mallard (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Muscovy Duck (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Graylag Goose (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Cockatiel (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "African Collared-Dove (Domestic type or Ringe") %>%
-  dplyr::filter(COMMON_NAME != "Budgerigar (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Red Junglefowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Helmeted Guineafowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Domestic goose sp. (Domestic type)") %>%
-  left_join(., clements, by="COMMON_NAME") %>%
-  dplyr::filter(family != "Stercorariidae (Skuas and Jaegers)") %>%
-  dplyr::filter(family != "Alcidae (Auks, Murres, and Puffins)") %>%
-  dplyr::filter(family != "Diomedeidae (Albatrosses)") %>%
-  dplyr::filter(family != "Oceanitidae (Southern Storm-Petrels)") %>%
-  dplyr::filter(family != "Hydrobatidae (Northern Storm-Petrels)") %>%
[…]

### 2019-04-09 R/predicting.R

@@ -1,58 +1,0 @@
-library(broom)
-library(tidyverse)
-library(arm)
-library(tidyr)
-
-load("Data/Modelling data/model_5_km_object.RData")
-
-oneday <- readRDS("Data/Sampling dates 2018/5 km/2018-12-31.RDS")
-
-
-add_dfbeta <- function(grid_size, date){
-  
-  # read in dataframe for a given date
-  df <- readRDS(paste0("Data/Sampling dates 2018/", grid_size, " km/", date, ".RDS"))
-  
-  # scale the df, according to how the
-  # dataframe for modelling was scaled
-  df2 <- df %>%
-    mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
-    mutate(s.duration=arm::rescale(duration)) %>%
-    mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
-    mutate(s.days_since=rescale(days_since)) %>%
-    mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
-    mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time)) %>%
-  # fill in the NA values for each of these variables
-  # necessary for predicting. Fill in with mean of that column
-    replace_na(list(
-      s.median_waiting_time=mean(.$s.median_waiting_time, na.rm=TRUE),
-      s.duration=mean(.$s.duration, na.rm=TRUE),
-      s.Number_of_days_sampled=mean(.$s.Number_of_days_sampled, na.rm=TRUE),
-      s.days_since=mean(.$s.days_since, na.rm=TRUE),
-      s.dist_km_nn=mean(.$s.dist_km_nn, na.rm=TRUE),
-      s.neighbor_waiting_time=mean(.$s.neighbor_waiting_time, na.rm=TRUE)
-      ))
-  
-  model <- readRDS(paste0("Data/Modelling data/model_", grid_size, "_km_object.RDS"))
-  
-  predicted_df <- augment(model, newdata = df2)
-  
-  saveRDS(predicted_df, paste0("Data/predicted_leverage/", grid_size, "_km/", date, ".RDS"))
-}
-
-
-dates_2018 <- seq.Date(as.Date("2018-01-01"), as.Date("2018-12-31"), by=1)
-
-
-lapply(dates_2018, function(x) {add_dfbeta(5, x)})
[…]

### 2019-04-09 R/prelim_mapping.R

@@ -1,57 +1,0 @@
-# prelim plotting
-
-library(ggplot2)
-library(sf)
-library(dplyr)
-library(tidyr)
-library(lubridate)
-
-# load data
-analysis_data <- readRDS("Data/Modelling data/analysis_data.RDS")
-
-load("Data/Spatial data/sydney_grids.RData")
-
-lev_results <- readRDS("Data/leverage_results.RDS")
-
-sites_grids_lookup <- readRDS("Data/sites_and_grids_lookup.RDS")
-
-
-sydney_grids %>%
-  st_set_geometry(NULL) %>%
-  rename(grid_id = id) %>%
-  left_join(., analysis_data, by="grid_id") %>%
-  group_by(grid_id) %>%
-  summarise(mean=mean(DATE_dfbetas, na.rm=TRUE),
-            min=min(DATE_dfbetas, na.rm=TRUE),
-            max=max(DATE_dfbetas, na.rm=TRUE),
-            quantile_.1=quantile(DATE_dfbetas, 0.1, na.rm=TRUE)) %>%
-  gather(., key="summary", value="value", mean, min, max, quantile_.1) %>%
-  left_join(., rename(sydney_grids, grid_id=id), by="grid_id") %>%
-  ggplot()+
-  geom_sf(aes(fill=value))+
-  facet_wrap(~summary)+
-  theme_classic()
-  
-
-sydney_grids %>%
-  st_set_geometry(NULL) %>%
-  rename(grid_id = id) %>%
-  left_join(., sites_grids_lookup, by="grid_id") %>%
-  left_join(., lev_results, by="LOCALITY_ID") %>%
-  mutate(YEAR = year(OBSERVATION_DATE)) %>%
-  group_by(grid_id, YEAR) %>%
-  summarise(mean=mean(DATE_dfbetas, na.rm=TRUE),
-            min=min(DATE_dfbetas, na.rm=TRUE),
-            max=max(DATE_dfbetas, na.rm=TRUE),
-            quantile_.1=quantile(DATE_dfbetas, 0.1, na.rm=TRUE)) %>%
-  gather(., key="summary", value="value", mean, min, max, quantile_.1) %>%
-  left_join(., rename(sydney_grids, grid_id=id), by="grid_id") %>%
-  ggplot()+
-  geom_sf(aes(fill=value))+
-  facet_wrap(YEAR~summary)+
-  theme_classic()
-
-
-
-
-

### 2019-04-09 R/random_extra_stuff.R

@@ -1,103 +1,0 @@
-# stuff to save for now
-unique_grids <- sydney_grids %>%
-  st_set_geometry(NULL) %>%
-  mutate(id=1:nrow(.)) %>%
-  mutate(id = as.character(as.integer(.$id)))
-
-grid_points_sf <- st_as_sf(unique_grids, coords=c("lon", "lat"), crs=st_crs(sydney_grids))
-
-nearest_df <- as.data.frame(t(as.data.frame(st_nn(grid_points_sf, sydney_grids, k=9))))
-
-nearest_df2 <- nearest_df %>%
-  rename(grid_id = V1) %>%
-  rename(n1 = V2) %>%
-  rename(n2 = V3) %>%
-  rename(n3 = V4) %>%
-  rename(n4 = V5) %>%
-  rename(n5 = V6) %>%
-  rename(n6 = V7) %>%
-  rename(n7 = V8) %>%
-  rename(n8 = V9)
-
-
-# assess how many of the nearest neighbors have been sampled
-nearest_sampled <- nearest_df2 %>%
-  left_join(., rename(sampled_or_not, n1=grid_id), by="n1") %>%
-  rename(sampled_n1 = sampled) %>%
-  dplyr::select(-n1) %>%
-  left_join(., rename(sampled_or_not, n2=grid_id), by="n2") %>%
-  rename(sampled_n2 = sampled) %>%
-  dplyr::select(-n2) %>%
-  left_join(., rename(sampled_or_not, n3=grid_id), by="n3") %>%
-  rename(sampled_n3 = sampled) %>%
-  dplyr::select(-n3) %>%
-  left_join(., rename(sampled_or_not, n4=grid_id), by="n4") %>%
-  rename(sampled_n4 = sampled) %>%
-  dplyr::select(-n4) %>%
-  left_join(., rename(sampled_or_not, n5=grid_id), by="n5") %>%
-  rename(sampled_n5 = sampled) %>%
-  dplyr::select(-n5) %>%
-  left_join(., rename(sampled_or_not, n6=grid_id), by="n6") %>%
-  rename(sampled_n6 = sampled) %>%
-  dplyr::select(-n6) %>%
-  left_join(., rename(sampled_or_not, n7=grid_id), by="n7") %>%
-  rename(sampled_n7 = sampled) %>%
-  dplyr::select(-n7) %>%
[…]

### 2019-04-09 R/some_plots_of_leverage_and_ebird_variables.R

@@ -1,32 +1,0 @@
-# AN R script to visualize how leverage 
-# behaves with a suite of ebird specific traits
-
-
-library(ggplot2)
-library(dplyr)
-
-lev <- readRDS("Data/leverage_results.RDS")
-
-# date versus leverage
-ggplot(lev, aes(x=OBSERVATION_DATE, y=DATE_dfbetas))+
-  geom_point()+
-  geom_smooth(method="lm")+
-  theme_bw()
-
-# cook's distance versus dfbeta
-ggplot(lev, aes(x=log(DATE_dfbetas), y=log(cooks_dist)))+
-  geom_point()+
-  geom_smooth(method="lm")+
-  theme_bw()
-
-# number of species versus dfbeta
-ggplot(lev, aes(x=Number_species, y=DATE_dfbetas))+
-  geom_point()+
-  geom_smooth(method="lm")+
-  theme_bw()
-
-# duration on a checklist versus dfbeta
-ggplot(lev, aes(x=DURATION_MINUTES, y=DATE_dfbetas))+
-  geom_point()+
-  geom_smooth(method="gam")+
-  theme_bw()

### 2019-04-09 R/species_pop_trend_example.R

@@ -1,411 +1,0 @@
-# An example species of a population trend model in time
-# packages
-library(dplyr)
-library(tidyr)
-library(lubridate)
-library(ggplot2)
-library(mgcv)
-library(lme4)
-library(car)
-library(readr)
-
-
-# load data
-load("Data/Greater_Sydney_data.RData")
-
-clements <- read_csv("Data/Clements-Checklist-v2018-August-2018.csv") %>%
-  rename(COMMON_NAME = `English name`)
-
-# filter the dataset
-# remove any incomplete checklists
-GS_observations <- GS_observations %>%
-  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
-  dplyr::filter(CATEGORY %in% c("species", "issf", "domestic")) %>%
-  dplyr::filter(COMMON_NAME != "Indian Peafowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Mallard (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Muscovy Duck (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Graylag Goose (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Cockatiel (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "African Collared-Dove (Domestic type or Ringe") %>%
-  dplyr::filter(COMMON_NAME != "Budgerigar (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Red Junglefowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Helmeted Guineafowl (Domestic type)") %>%
-  dplyr::filter(COMMON_NAME != "Domestic goose sp. (Domestic type)") %>%
-  left_join(., clements, by="COMMON_NAME") %>%
-  dplyr::filter(family != "Stercorariidae (Skuas and Jaegers)") %>%
-  dplyr::filter(family != "Alcidae (Auks, Murres, and Puffins)") %>%
-  dplyr::filter(family != "Diomedeidae (Albatrosses)") %>%
-  dplyr::filter(family != "Oceanitidae (Southern Storm-Petrels)") %>%
-  dplyr::filter(family != "Hydrobatidae (Northern Storm-Petrels)") %>%
[…]

### 2019-04-09 R/tutorial_for_citizen_science_essay.R
no idea what I'm doing!

@@ -1,0 +1,374 @@
+# An R script to make stuff for a tutorial of sorts on how things 
+# are calculated to parameterize the value of a BSE
+
+# packages
+library(sf)
+library(raster)
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(lubridate)
+library(data.table)
+library(geosphere)
+library(BBmisc)
+library(RANN)
+library(ggsn)
+library(readr)
+
+
+# load shapefile
+sydney <- st_read("Data/Spatial data/greater_sydney_shape/sydney.shp")
+
+# read data in
+load("Data/Sampling data/NSW_birds.RData")
+
+
+# select ten 'locations/sites'
+# this is done by just browsing on the eBird website
+example_sites <- c("L945869", "L2557723", "L8083126", "L1030678", "L915566", 
+                   "L1300136", "L5076722", "L2444301", "L3007885", "L5146094")
+
+
+# get data out for those ten sites
+example_data <- NSW_ebird %>%
+  dplyr::filter(PROTOCOL_TYPE != "Incidental") %>%
+  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
+  dplyr::filter(OBSERVATION_DATE >= "2010-01-01") %>%
+  dplyr::filter(LOCALITY_ID %in% example_sites)
+
+# Make a plot of those ten locations in space
+sites_coords <- NSW_ebird %>%
+  dplyr::filter(LOCALITY_ID %in% example_sites) %>%
+  dplyr::select(LOCALITY_ID, lat, lng) %>%
+  distinct()
+
+ggplot()+
+  geom_sf(data=sydney, fill="lightblue")+
+  geom_point(data=sites_coords, aes(x=lng, y=lat), color="firebrick3", size=2.3)+
+  coord_sf(xlim = c(151.05, 151.3), ylim = c(-33.97, -33.77))+
+  xlab("")+
+  ylab("")+
+  theme_bw()+
+  theme(panel.grid.major = element_line(colour = 'transparent'))
+
+ggsave(filename = "C:/Users/CTC/Desktop/citizen_science_value/Submissions/PLoS Biology/Revision 1/Appendix/inset.png",
+       height=4, width=4, units="in")
+
+ggplot()+
+  geom_sf(data=sydney, fill="lightblue")+
+  geom_point(data=sites_coords, aes(x=lng, y=lat), color="firebrick3", size=2.3)+
+  xlab("")+
[…]

### 2019-04-09 R/tutorial_for_citizen_science_essay.R
Revert "no idea what I'm doing!"
This reverts commit 06f849a313b742515dfd3b937d8c18b4efd91e8c.

@@ -1,374 +1,0 @@
-# An R script to make stuff for a tutorial of sorts on how things 
-# are calculated to parameterize the value of a BSE
-
-# packages
-library(sf)
-library(raster)
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(lubridate)
-library(data.table)
-library(geosphere)
-library(BBmisc)
-library(RANN)
-library(ggsn)
-library(readr)
-
-
-# load shapefile
-sydney <- st_read("Data/Spatial data/greater_sydney_shape/sydney.shp")
-
-# read data in
-load("Data/Sampling data/NSW_birds.RData")
-
-
-# select ten 'locations/sites'
-# this is done by just browsing on the eBird website
-example_sites <- c("L945869", "L2557723", "L8083126", "L1030678", "L915566", 
-                   "L1300136", "L5076722", "L2444301", "L3007885", "L5146094")
-
-
-# get data out for those ten sites
-example_data <- NSW_ebird %>%
-  dplyr::filter(PROTOCOL_TYPE != "Incidental") %>%
-  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
-  dplyr::filter(OBSERVATION_DATE >= "2010-01-01") %>%
-  dplyr::filter(LOCALITY_ID %in% example_sites)
-
-# Make a plot of those ten locations in space
-sites_coords <- NSW_ebird %>%
-  dplyr::filter(LOCALITY_ID %in% example_sites) %>%
-  dplyr::select(LOCALITY_ID, lat, lng) %>%
-  distinct()
-
-ggplot()+
-  geom_sf(data=sydney, fill="lightblue")+
-  geom_point(data=sites_coords, aes(x=lng, y=lat), color="firebrick3", size=2.3)+
-  coord_sf(xlim = c(151.05, 151.3), ylim = c(-33.97, -33.77))+
-  xlab("")+
-  ylab("")+
-  theme_bw()+
-  theme(panel.grid.major = element_line(colour = 'transparent'))
-
-ggsave(filename = "C:/Users/CTC/Desktop/citizen_science_value/Submissions/PLoS Biology/Revision 1/Appendix/inset.png",
-       height=4, width=4, units="in")
-
-ggplot()+
-  geom_sf(data=sydney, fill="lightblue")+
-  geom_point(data=sites_coords, aes(x=lng, y=lat), color="firebrick3", size=2.3)+
-  xlab("")+
[…]

### 2019-04-09 optimize_citi_sci_obs.Rmd
no message

---
title: "Optimizing citizen science observation sampling"

author: |
  | Corey T. Callaghan^[Corresponding author: c.callaghan@unsw.edu.au] $^1$, Jodi J. L. Rowley $^1$$^,$$^2$, Richard E. Major $^1$$^,$$^2$, Alistair G. B. Poore $^3$, William K. Cornwell $^1$$^,$$^3$
  | $^1$Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia; $^2$Australian Museum Research Institute, Sydney, NSW, Australia; $^3$Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia
date: ""
output: word_document
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/plos.csl
bibliography: references.bib
---

\newpage

# ABSTRACT

*Keywords*: citizen science, biodiversity sampling, spatial sampling, temporal sampling

\newpage
# Introduction
Assessing biodiversity trends in space and time is essential for conservation  [@harrison2014assessing; @wilson2011modelling; @mcmahon2011improving; @honrado2016fostering; @dornelas2013quantifying]. Unsurprisingly, trend detection is best estimated when using data from well-designed surveys in space and time [@harrison2014assessing; @vellend2017estimates; @kery2009trend], that are long-term [@lindenmayer2012value]. But scientific funding for long-term ecological research is rapidly disappearing [REFS???]. Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local [@callaghan2015efficacy; @theobald2015global; @sullivan2017using; @loss2015linking], regional [@barlow2015citizen; @fox2011new], and global scales [@chandler2017contribution; @pocock2018vision; @cooper2014invisible].

Citizen science projects fall along a continuum of the level of associated structure [@kelling2019using; @welvaert2016citizen], ranging from unstructured (e.g., iNaturalist) to structured (U.K. Butterfly Monitoring Scheme). All citizen science projects have associated biases, including spatial and temporal biases [@boakes2010distorted], and different levels of observers' skills [@kelling2015can]. But these biases can generally be accounted for, statistically [@isaac2014statistics; @robinson2018correcting], by filtering or subsetting the data [@wiggins2011conservation], pooling multiple data sources [@fithian2015bias], or machine learning and hierarchical clustering techniques [@hochachka2012data; @kelling2015taking]. Regardless of biases, estimating trends is best done with structured projects [e.g. @fox2011new], but unstructured and semistructured projects are increasingly harnessed for trend detection [@walker2017using; @kery2009trend; @kery2010site; @horns2018using; @van2013occupancy; @pagel2014quantifying].

The number of ecological and environmental monitoring citizen science projects is increasing [@pocock2017diversity; @theobald2015global], highlighting the potential of these data in the future of ecology, conservation, and resource management [@pocock2018vision; @mckinley2017citizen]. But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets [@parrish2018exposing].

Despite the potential of citizen science for the future of conservation ecology [@pocock2018vision; @silvertown2009new; @soroye2018opportunistic], can it be better? Can methods be developed which optimize sampling for the data-collection phase, decreasing the statistical error associated with population trends? We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which *should* be sampled on any given day. This approach relies on statistical leverage, assigning every citizen science observation a measure of influence that influences a statistical model. Here, we illustrate our vision using a popular citizen science project --- eBird [@sullivan2009ebird].

# Methods
## Study region
We illustrate our concepts throughout the Greater Sydney Region, and we gridded the region into 411 grids, all of which are ~ km. We used the R statistical environment to carry out all analyses.

## eBird data
We downloaded the eBird basic dataset (VERSION), and subsetted the data between January 1st, 2010 to December 31st, 2018. eBird collects data in the form of 'checklists' --- a list of all species identified (audibly or visually) for a given spatiotemporal coordinates. eBird relies on an extensive network of regional reviewers who are local experts of the avifauna [REF]

## Trend detection
For illustration purposes, we fitted a simple generalized linear model, based on presence/absence for each of 274 species --- those of which had > 20 observations in the region throughout the study period. The GLM took the form of: XXXXXX. County was included to add a spatial representation throughout the Greater Sydney Region, while we included the length of the list in the model as an offset to account for the effort associated with each individual eBird checklist.

We acknowledge that a number of different models can be fit which assesses trend detection [REFS], and biases can be accounted for in a variety of ways [REFS]. These models were fitted for the period between 2010-2018.

### Statistical leverage
Statistical leverage measures the influence of a particular observation on the independent variable [REFS]. In other words, it is a measure of how much an observation influences the results of an observation. In our instance, because we had multiple predictor variables in our GLMs, we used dfBeta as a measure of statistical leverage. dfBeta can be calculated for a given parameter in a regression model.

For each species (N=274) we calculated the dfBeta for each of the 29,055 observations used for each model. Then for every checklist assessed in the model, we took the sum of the absolute value of the dfBetas for that checklist as a measure of a checklist's leverage in understanding species' trends throughout the Greater Sydney Region.

## Parameter estimation
For each day in 2018 (N=365), for each grid, we dynamically calculated the following parameters, based on sampling: (1) whether a grid cell was sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date. These are some hypothesized parameters which we felt would influence the relative value of a given sampling event.

We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids. We ran a linear regression to parameterize what parameters had the greatest impact on the leverage of a given observation.

## Calculating expected leverage

# Results
A total of XXX observations (i.e., eBird checklists) were used to produce trend estimates for each of XXX species.

# Discussion
- Looking forward but using the past
- We only looked at trend detection, but many other models can be used in this process (e.g., species distribution models)
- Other potentials for this general approach (identifying outliers in the dataset or misidentified species)

\newpage
# References

### 2019-04-09 R/build_df_for_parameterizing_model.R

@@ -1,0 +1,232 @@
+### Create a dataframe for which modelling against the formula
+
+
+# packages
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(sf)
+library(data.table)
+library(geosphere)
+library(BBmisc)
+library(RANN)
+
+
+# load data
+load("Data/Greater_Sydney_data.RData")
+
+load("Data/Sampling data/NSW_birds.RData")
+
+load("Data/Spatial data/sydney_grids.RData")
+
+
+
+# assign each locality (and thereby sampling event identifier) 
+# a grid cell throughout Sydney
+sydney_grids <- sydney_grids %>%
+  rename(grid_id = id)
+
+# get list of unique sites
+sites <- GS_observations %>%
+  dplyr::select(LOCALITY_ID, LATITUDE, LONGITUDE) %>%
+  distinct()
+
+
+# convert points from FrogID into sf points
+# so that they can be used with the polygons
+crs <- "+proj=longlat +ellps=GRS80 +no_defs"
+
+sites_sf <- st_as_sf(sites, coords = c("LONGITUDE", "LATITUDE"), 
+                      crs = crs, agr = "constant")
+
+# use intersect to return a list of points and the associated
+# polygon they belong to - uses integer ids for rows and columns
+# note that it looks like 29 points aren't assigned to a LGA
+# not worried about this for now, but something to come back to in the future
+assigned_points <- as.data.frame(st_intersects(sites_sf, sydney_grids))
+
+# join the original dataframe with the intersected df
+# then remove the row and col ids
+sites_and_grids <- sites %>%
+  mutate(row.id = 1:nrow(.)) %>%
+  left_join(assigned_points, by="row.id") %>%
+  rename(grid_id = col.id) %>%
+  dplyr::select(-row.id)
+
+
+test <- sydney_grids %>%
+  left_join(., sites_and_grids, by="grid_id") %>%
+  left_join(., GS_observations, by="LOCALITY_ID") %>%
+  group_by(grid_id) %>%
+  summarise(N=length(unique(SAMPLING_EVENT_IDENTIFIER))) 
+
+
+
+
+
+date_specific_summary <- function(forecast_date) {
+
+# do it for a single date first
[…]

### 2019-04-09 R/build_df_for_parameterizing_model_for_grids.R

@@ -1,0 +1,331 @@
+### Create a dataframe for which modelling against the formula can be done
+### This means that for everyday throughout 2018 I will calculate what grids have
+### been sampled and the various aspects that might be useful in
+### calculating where to go sample next
+### this dataframe will then be put together with the leverage of every checklist
+### to see how well people are sampling
+### will do this for every day in 2018
+### so that the function is passed a date and then lapply over a list of dates
+
+
+# packages
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(sf)
+library(data.table)
+library(geosphere)
+library(BBmisc)
+library(RANN)
+library(nngeo)
+
+
+# load data
+load("Data/Greater_Sydney_data.RData")
+load("Data/Sampling data/NSW_birds.RData")
+load("Data/Spatial data/sydney_grids_1_km.RData")
+load("Data/Spatial data/sydney_grids_2_km.RData")
+load("Data/Spatial data/sydney_grids_5_km.RData")
+load("Data/Spatial data/sydney_grids_10_km.RData")
+load("Data/Spatial data/sydney_grids_25_km.RData")
+load("Data/Spatial data/sydney_grids_50_km.RData")
+
+
+
+# assign each locality (and thereby sampling event identifier) 
+# a grid cell throughout Sydney
+# will be necessary for merging back later
+# because the calculations will be on a grid scale
+# get list of unique sites (LOCALITY_ID)
+sites <- GS_observations %>%
+  dplyr::select(LOCALITY_ID, LATITUDE, LONGITUDE) %>%
+  distinct()
+
+
+# convert points into sf points
+# so that they can be used with the polygons
+crs <- 4326
+
+sites_sf <- st_as_sf(sites, coords = c("LONGITUDE", "LATITUDE"), 
+                      crs = crs, agr = "constant")
+
+
+date_specific_summary <- function(forecast_date, dataset, grid_size) {
+
+  
+# use intersect to return a list of points and the associated
+# polygon they belong to - uses integer ids for rows and columns
+assigned_points <- as.data.frame(st_intersects(sites_sf, dataset))
+
+# join the original dataframe with the intersected df
[…]

### 2019-04-09 R/create_model_df.R

@@ -1,0 +1,182 @@
+# An R script which reads in each of the day's sampling
+# summary, for each of the different grid cell sizes
+# and makes dataframes.
+# then combines those dataframes into the 'parameters' df
+
+library(dplyr)
+library(tidyr)
+
+
+
+# get a dataframe for modelling 
+# for each of the different grid cell sizes
+
+# 50 km first
+setwd("Data/Sampling dates 2018/50 km")
+
+file_list <- list.files()
+
+
+for (file in file_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("dataset")){
+    dataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, append to it
+  if (exists("dataset")){
+    temp_dataset <-readRDS(file)
+    dataset<-rbind(dataset, temp_dataset)
+    rm(temp_dataset)
+  }
+  
+}
+
+dataset1 <- dataset %>%
+  distinct() %>%
+  mutate(grid_size=50)
+
+rm(dataset)
+
+setwd('..')
+setwd('..')
+setwd('..')
+
+# 25 km
+setwd("Data/Sampling dates 2018/25 km")
+
+file_list <- list.files()
+
+
+for (file in file_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("dataset")){
+    dataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, append to it
+  if (exists("dataset")){
+    temp_dataset <-readRDS(file)
+    dataset<-rbind(dataset, temp_dataset)
+    rm(temp_dataset)
+  }
+  
+}
+
+dataset2 <- dataset %>%
+  distinct() %>%
+  mutate(grid_size=25)
+
+rm(dataset)
+
+setwd('..')
+setwd('..')
+setwd('..')
+
+# 10 km
+setwd("Data/Sampling dates 2018/10 km")
+
+file_list <- list.files()
+
+
+for (file in file_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("dataset")){
+    dataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, append to it
+  if (exists("dataset")){
+    temp_dataset <-readRDS(file)
+    dataset<-rbind(dataset, temp_dataset)
+    rm(temp_dataset)
+  }
+  
+}
+
+dataset3 <- dataset %>%
[…]

### 2019-04-09 R/create_raster_of_sydney.R

@@ -1,0 +1,78 @@
+# create a raster of the study area
+
+library(sf)
+library(dplyr)
+
+
+# load shapefile
+sydney <- st_read("Data/Spatial data/greater_sydney_shape/sydney.shp") %>%
+  st_transform(32756)
+
+
+# function to make a grid over Sydney region
+# grid size is defined in meters
+get_gridded_region_function <- function(grid_size) {
+  
+  sydney_grids <- sydney %>%
+    st_make_grid(., cellsize=c(grid_size, grid_size),
+                 crs=32756, what='polygons') %>%
+    st_intersection(sydney) %>%
+    st_cast("MULTIPOLYGON") %>%
+    st_sf() %>%
+    mutate(id = row_number())
+  
+  
+  centroids <- st_centroid(sydney_grids)
+  centroids <- st_transform(centroids, 4326)
+  sydney_grids <- st_transform(sydney_grids, 4326)
+  
+  coords <- do.call(rbind, st_geometry(centroids)) %>%
+    as_tibble() %>%
+    setNames(c("lon", "lat"))
+  
+  sydney_grids <- bind_cols(sydney_grids, coords) %>%
+    rename(grid_id = id)
+  
+  return(sydney_grids)
+  
+}
+
+# 50 km grids
+sydney_grids_50_km <- get_gridded_region_function(grid_size=50000)
+
+save(sydney_grids_50_km, file = "Data/Spatial data/sydney_grids_50_km.RData")
+
+
+# 25 km grids
+sydney_grids_25_km <- get_gridded_region_function(grid_size=25000)
+
+save(sydney_grids_25_km, file = "Data/Spatial data/sydney_grids_25_km.RData")
+
+
+# 10 km grids
+sydney_grids_10_km <- get_gridded_region_function(grid_size=10000)
+
+save(sydney_grids_10_km, file = "Data/Spatial data/sydney_grids_10_km.RData")
+
+
+# 5 km grids
+sydney_grids_5_km <- get_gridded_region_function(grid_size=5000)
+
+save(sydney_grids_5_km, file = "Data/Spatial data/sydney_grids_5_km.RData")
+
+
+# 2 km grids
+sydney_grids_2_km <- get_gridded_region_function(grid_size=2000)
+
+save(sydney_grids_2_km, file = "Data/Spatial data/sydney_grids_2_km.RData")
+
+
+# 1 km grids
[…]

### 2019-04-09 R/data_frame_testing.R

@@ -1,0 +1,12 @@
+library(tidyverse)
+
+lev<-readRDS("Data/Modelling data/leverage_results.RDS")
+param<-readRDS("Data/Modelling data/parameter_dataset.RDS")
+
+
+all<-left_join(lev,param)
+
+#should be a lot of 4's 
+sort(table(all$SAMPLING_EVENT_IDENTIFIER),decreasing = T)
+
+

### 2019-04-09 R/get_all_observations_from_greater_sydney.R

@@ -1,0 +1,77 @@
+# An R script to get all eBird data subsetted to the greater sydney region
+# also filtered to get rid of 'bad' checklists
+# it takes a long time for these filtering of the MariaDB to be done
+# so go get a coffee or something (or a beer!!)
+
+library(sf)
+library(dbplyr)
+library(dplyr, quietly = T, warn.conflicts = F)
+library(RMariaDB)
+get_con  <- function () {
+  dbConnect(RMariaDB::MariaDB(), host = 'KESTREL', dbname='ebird',user = Sys.getenv('userid'), password = Sys.getenv('pwd'))
+} 
+
+sites <- tbl(con, "sites")
+samples <- tbl(con, "samples")
+species <- tbl(con, "species")
+checklists <- tbl(con, "checklists")
+
+NSW_species <- sites %>% 
+  filter(STATE_PROVINCE == "New South Wales") %>% 
+  left_join(., checklists, by="LOCALITY_ID") %>% 
+  inner_join(., samples, by="SAMPLING_EVENT_IDENTIFIER") %>% 
+  collect(n=Inf)
+
+NSW_sites <- sites %>%
+  filter(STATE_PROVINCE == "New South Wales") %>%
+  collect(n=Inf)
+
+species_all <- species %>%
+  select(TAXONOMIC_ORDER, CATEGORY, COMMON_NAME, SCIENTIFIC_NAME) %>%
+  collect(n=Inf)
+
+## assign each observation to a Greater Capital City Statistical Areas (GCCSA)
+## This step requires download of shapefile from
+## here http://www.abs.gov.au/AUSSTATS/abs@.nsf/DetailsPage/1270.0.55.001July%202016?OpenDocument )
+## we then read in the shapefile and assign each point to a particular GCCSA
+## this will be used for later visualizations and analyses
+
+
+# read in the shapefile
+aus_gccsa <- st_read("Data/Spatial data/gccsa_2016_aust_shape/GCCSA_2016_AUST.shp")
+
+# turn observations into sf object
+# only for unique lists
+points_sf <- st_as_sf(NSW_sites, coords = c("LONGITUDE", "LATITUDE"), crs=st_crs(aus_gccsa)) 
+
+# create metadata df
+# used to link the dfs back together later
+gccsa_metadata <- data.frame(gccsa_name=aus_gccsa[[2]], col.id=1:34)
+
[…]

### 2019-04-09 R/initial_script_for_formula.R

@@ -1,0 +1,458 @@
+# An initial R script, attempting to develop a 'formula'
+# which can provide value to a given site
+# relative to other sites
+
+
+
+# packages
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(sf)
+library(data.table)
+library(geosphere)
+library(BBmisc)
+library(RANN)
+
+# read data in
+load("Data/Sampling data/NSW_birds.RData")
+
+
+# select ten 'locations/sites'
+# this is done by just browsing on th eBird 
+# website for now
+# eventually
+# if we get a formula that 'works'
+# then we will want to come back to how to define a site
+example_sites <- c("L945869", "L2557723", "L8083126", "L1030678", "L915566", 
+                   "L1300136", "L5076722", "L2444301", "L3007885", "L5146094")
+
+
+# get data out for those ten sites
+example_data <- NSW_ebird %>%
+  dplyr::filter(PROTOCOL_TYPE != "Incidental") %>%
+  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
+  dplyr::filter(OBSERVATION_DATE >= "2010-01-01") %>%
+  dplyr::filter(LOCALITY_ID %in% example_sites)
+
+
+# now need to calculate the following variables for
+# each site
+# 1.) Whether a site was sampled or not
+# 2.) distance from the nearest sampled site (nearest neighbor)
+# 3.) The median waiting time between checklists at the site
+# 4.) days between first and last observation
+# 5.) The median waiting time at the nearest neighbor
+# 6.) Days since last sampling event
+
+
+# 1.) Whether a site was sampled or not
+# all sites
+all_sites <- NSW_ebird %>%
+  dplyr::filter(LOCALITY_ID %in% example_sites) %>%
+  dplyr::select(LOCALITY_ID) %>%
+  distinct() 
+
+# sampled sites
+sampled_sites <- example_data %>%
+  dplyr::select(LOCALITY_ID) %>%
+  distinct() %>%
+  mutate(sampled = "yes")
+
+sampled_or_not <- all_sites %>%
+  left_join(., sampled_sites, by="LOCALITY_ID") %>%
+  replace_na(list(sampled="no"))
+
+
+# 2.) distance from the nearest sampled site
+# not sure how well this approach will scale for a large
+# number of 'sites'
+# but it works for now
[…]

### 2019-04-09 R/modelling.R

@@ -1,0 +1,317 @@
+### Modelling script
+
+# packages
+library(dplyr)
+library(ggplot2)
+library(BBmisc)
+library(ggcorrplot)
+library(corrplot)
+library(lme4)
+library(standardize)
+library(tibble)
+library(visreg)
+library(arm)
+
+# load data
+lev <- readRDS("Data/Modelling data/leverage_results.RDS")
+
+params <- readRDS("Data/Modelling data/parameter_dataset.RDS")
+
+analysis_data <- params %>%
+  left_join(., lev) %>%
+  mutate(
+    sampled_numeric=case_when(
+    sampled=="yes" ~ 1,
+    sampled=="no" ~ 0)
+    )
+
+# make df for modelling
+hist(analysis_data$DATE_dfbetas, breaks=30)
+
+hist(log(analysis_data$DATE_dfbetas), breaks=30)
+
+# rescale variables
+analysis_data <- analysis_data %>%
+  group_by(grid_size) %>%
+  mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
+  mutate(s.duration=arm::rescale(duration)) %>%
+  mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
+  mutate(s.days_since=rescale(days_since)) %>%
+  mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
+  mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time))
+
+# look at correlation among predictor variables
+cor_object5 <- analysis_data %>%
+  dplyr::filter(grid_size==5) %>%
+  dplyr::select(s.median_waiting_time, s.duration, s.Number_of_days_sampled, 
+                s.days_since, s.dist_km_nn, s.neighbor_waiting_time) %>%
+  ungroup()
+
+ggcorrplot(cor(na.omit(cor_object5[2:7])))+
+  ggtitle("5 km grid size")
+
+# look at correlation among predictor variables
+cor_object10 <- analysis_data %>%
+  dplyr::filter(grid_size==10) %>%
+  dplyr::select(s.median_waiting_time, s.duration, s.Number_of_days_sampled, 
+                s.days_since, s.dist_km_nn, s.neighbor_waiting_time) %>%
+  ungroup()
+
+ggcorrplot(cor(na.omit(cor_object10[2:7])))+
+  ggtitle("10 km grid size")
+
+
+# look at correlation among predictor variables
[…]

### 2019-04-09 R/plot_leverage.R

@@ -1,0 +1,122 @@
+## An R script to plot predicted leverage
+## Will do it for 5 km grid cell size first
+
+library(dplyr)
+library(sf)
+library(ggplot2)
+library(wesanderson)
+
+# 5 km first
+setwd("Data/predicted_leverage/5_km")
+
+file_list <- list.files()
+
+
+for (file in file_list){
+  
+  # if the merged dataset doesn't exist, create it
+  if (!exists("dataset")){
+    dataset <- readRDS(file)
+  }
+  
+  # if the merged dataset does exist, append to it
+  if (exists("dataset")){
+    temp_dataset <-readRDS(file)
+    dataset<-rbind(dataset, temp_dataset)
+    rm(temp_dataset)
+  }
+  
+}
+
+setwd('..')
+setwd('..')
+setwd('..')
+
+load("Data/Spatial data/sydney_grids_5_km.RData")
+
+
+leverage_5 <- dataset %>%
+  distinct() %>%
+  mutate(Value=exp(.fitted)) %>%
+  group_by(grid_id, dynamic_date) %>%
+  sample_n(1) %>%
+  ungroup() 
+
+
+
+df <- leverage_5 %>%
+  dplyr::filter(dynamic_date==date)
+
+map_df <- sydney_grids_5_km %>%
+  right_join(., df, by="grid_id")
+
+plasma_pal <- c("red", viridis::plasma(n = 6))
+
+ggplot()+
+  geom_sf(data=map_df, aes(fill=Value))+
+  theme_bw()+
+  theme(panel.grid.major=element_line(color="transparent"))+
+  theme(axis.text=element_blank())+
+  theme(axis.ticks=element_blank())+
+  scale_fill_gradientn(colours=pal, name="Value", trans="log")
+  
+# potential figure for paper
+example_dates <- c("2018-01-01", "2018-03-01", "2018-05-01",
+                   "2018-07-01", "2018-09-01", "2018-11-01")
+
+df <- leverage_5 %>%
+  dplyr::filter(as.character(as.Date(dynamic_date)) %in% example_dates)
+
+map_df <- sydney_grids_5_km %>%
+  right_join(., df, by="grid_id") %>%
+  st_simplify()
+
+
+ggplot()+
+  geom_sf(data=map_df, aes(fill=Value))+
+  theme_bw()+
+  theme(panel.grid.major=element_line(color="transparent"))+
+  theme(axis.text=element_blank())+
[…]

### 2019-04-09 R/population_trend_model.R

@@ -1,0 +1,262 @@
+## A function to apply for every species to assess 'population trends'
+
+# packages
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(ggplot2)
+library(mgcv)
+library(lme4)
+library(car)
+library(readr)
+
+
+# load data
+load("Data/Greater_Sydney_data.RData")
+
+clements <- read_csv("Data/Clements-Checklist-v2018-August-2018.csv") %>%
+  rename(COMMON_NAME = `English name`)
+
+# filter the dataset
+# remove any incomplete checklists
+GS_observations <- GS_observations %>%
+  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
+  dplyr::filter(CATEGORY %in% c("species", "issf", "domestic")) %>%
+  dplyr::filter(COMMON_NAME != "Indian Peafowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Mallard (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Muscovy Duck (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Graylag Goose (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Cockatiel (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "African Collared-Dove (Domestic type or Ringe") %>%
+  dplyr::filter(COMMON_NAME != "Budgerigar (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Red Junglefowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Helmeted Guineafowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Domestic goose sp. (Domestic type)") %>%
+  left_join(., clements, by="COMMON_NAME") %>%
+  dplyr::filter(family != "Stercorariidae (Skuas and Jaegers)") %>%
+  dplyr::filter(family != "Alcidae (Auks, Murres, and Puffins)") %>%
+  dplyr::filter(family != "Diomedeidae (Albatrosses)") %>%
+  dplyr::filter(family != "Oceanitidae (Southern Storm-Petrels)") %>%
+  dplyr::filter(family != "Hydrobatidae (Northern Storm-Petrels)") %>%
[…]

### 2019-04-09 R/predicting.R

@@ -1,0 +1,58 @@
+library(broom)
+library(tidyverse)
+library(arm)
+library(tidyr)
+
+load("Data/Modelling data/model_5_km_object.RData")
+
+oneday <- readRDS("Data/Sampling dates 2018/5 km/2018-12-31.RDS")
+
+
+add_dfbeta <- function(grid_size, date){
+  
+  # read in dataframe for a given date
+  df <- readRDS(paste0("Data/Sampling dates 2018/", grid_size, " km/", date, ".RDS"))
+  
+  # scale the df, according to how the
+  # dataframe for modelling was scaled
+  df2 <- df %>%
+    mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
+    mutate(s.duration=arm::rescale(duration)) %>%
+    mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
+    mutate(s.days_since=rescale(days_since)) %>%
+    mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
+    mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time)) %>%
+  # fill in the NA values for each of these variables
+  # necessary for predicting. Fill in with mean of that column
+    replace_na(list(
+      s.median_waiting_time=mean(.$s.median_waiting_time, na.rm=TRUE),
+      s.duration=mean(.$s.duration, na.rm=TRUE),
+      s.Number_of_days_sampled=mean(.$s.Number_of_days_sampled, na.rm=TRUE),
+      s.days_since=mean(.$s.days_since, na.rm=TRUE),
+      s.dist_km_nn=mean(.$s.dist_km_nn, na.rm=TRUE),
+      s.neighbor_waiting_time=mean(.$s.neighbor_waiting_time, na.rm=TRUE)
+      ))
+  
+  model <- readRDS(paste0("Data/Modelling data/model_", grid_size, "_km_object.RDS"))
+  
+  predicted_df <- augment(model, newdata = df2)
+  
+  saveRDS(predicted_df, paste0("Data/predicted_leverage/", grid_size, "_km/", date, ".RDS"))
+}
+
+
+dates_2018 <- seq.Date(as.Date("2018-01-01"), as.Date("2018-12-31"), by=1)
+
+
+lapply(dates_2018, function(x) {add_dfbeta(5, x)})
[…]

### 2019-04-09 R/prelim_mapping.R

@@ -1,0 +1,57 @@
+# prelim plotting
+
+library(ggplot2)
+library(sf)
+library(dplyr)
+library(tidyr)
+library(lubridate)
+
+# load data
+analysis_data <- readRDS("Data/Modelling data/analysis_data.RDS")
+
+load("Data/Spatial data/sydney_grids.RData")
+
+lev_results <- readRDS("Data/leverage_results.RDS")
+
+sites_grids_lookup <- readRDS("Data/sites_and_grids_lookup.RDS")
+
+
+sydney_grids %>%
+  st_set_geometry(NULL) %>%
+  rename(grid_id = id) %>%
+  left_join(., analysis_data, by="grid_id") %>%
+  group_by(grid_id) %>%
+  summarise(mean=mean(DATE_dfbetas, na.rm=TRUE),
+            min=min(DATE_dfbetas, na.rm=TRUE),
+            max=max(DATE_dfbetas, na.rm=TRUE),
+            quantile_.1=quantile(DATE_dfbetas, 0.1, na.rm=TRUE)) %>%
+  gather(., key="summary", value="value", mean, min, max, quantile_.1) %>%
+  left_join(., rename(sydney_grids, grid_id=id), by="grid_id") %>%
+  ggplot()+
+  geom_sf(aes(fill=value))+
+  facet_wrap(~summary)+
+  theme_classic()
+  
+
+sydney_grids %>%
+  st_set_geometry(NULL) %>%
+  rename(grid_id = id) %>%
+  left_join(., sites_grids_lookup, by="grid_id") %>%
+  left_join(., lev_results, by="LOCALITY_ID") %>%
+  mutate(YEAR = year(OBSERVATION_DATE)) %>%
+  group_by(grid_id, YEAR) %>%
+  summarise(mean=mean(DATE_dfbetas, na.rm=TRUE),
+            min=min(DATE_dfbetas, na.rm=TRUE),
+            max=max(DATE_dfbetas, na.rm=TRUE),
+            quantile_.1=quantile(DATE_dfbetas, 0.1, na.rm=TRUE)) %>%
+  gather(., key="summary", value="value", mean, min, max, quantile_.1) %>%
+  left_join(., rename(sydney_grids, grid_id=id), by="grid_id") %>%
+  ggplot()+
+  geom_sf(aes(fill=value))+
+  facet_wrap(YEAR~summary)+
+  theme_classic()
+
+
+
+
+

### 2019-04-09 R/some_plots_of_leverage_and_ebird_variables.R

@@ -1,0 +1,32 @@
+# AN R script to visualize how leverage 
+# behaves with a suite of ebird specific traits
+
+
+library(ggplot2)
+library(dplyr)
+
+lev <- readRDS("Data/leverage_results.RDS")
+
+# date versus leverage
+ggplot(lev, aes(x=OBSERVATION_DATE, y=DATE_dfbetas))+
+  geom_point()+
+  geom_smooth(method="lm")+
+  theme_bw()
+
+# cook's distance versus dfbeta
+ggplot(lev, aes(x=log(DATE_dfbetas), y=log(cooks_dist)))+
+  geom_point()+
+  geom_smooth(method="lm")+
+  theme_bw()
+
+# number of species versus dfbeta
+ggplot(lev, aes(x=Number_species, y=DATE_dfbetas))+
+  geom_point()+
+  geom_smooth(method="lm")+
+  theme_bw()
+
+# duration on a checklist versus dfbeta
+ggplot(lev, aes(x=DURATION_MINUTES, y=DATE_dfbetas))+
+  geom_point()+
+  geom_smooth(method="gam")+
+  theme_bw()

### 2019-04-09 R/species_pop_trend_example.R

@@ -1,0 +1,411 @@
+# An example species of a population trend model in time
+# packages
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(ggplot2)
+library(mgcv)
+library(lme4)
+library(car)
+library(readr)
+
+
+# load data
+load("Data/Greater_Sydney_data.RData")
+
+clements <- read_csv("Data/Clements-Checklist-v2018-August-2018.csv") %>%
+  rename(COMMON_NAME = `English name`)
+
+# filter the dataset
+# remove any incomplete checklists
+GS_observations <- GS_observations %>%
+  dplyr::filter(ALL_SPECIES_REPORTED == 1) %>%
+  dplyr::filter(CATEGORY %in% c("species", "issf", "domestic")) %>%
+  dplyr::filter(COMMON_NAME != "Indian Peafowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Mallard (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Muscovy Duck (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Graylag Goose (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Cockatiel (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "African Collared-Dove (Domestic type or Ringe") %>%
+  dplyr::filter(COMMON_NAME != "Budgerigar (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Red Junglefowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Helmeted Guineafowl (Domestic type)") %>%
+  dplyr::filter(COMMON_NAME != "Domestic goose sp. (Domestic type)") %>%
+  left_join(., clements, by="COMMON_NAME") %>%
+  dplyr::filter(family != "Stercorariidae (Skuas and Jaegers)") %>%
+  dplyr::filter(family != "Alcidae (Auks, Murres, and Puffins)") %>%
+  dplyr::filter(family != "Diomedeidae (Albatrosses)") %>%
+  dplyr::filter(family != "Oceanitidae (Southern Storm-Petrels)") %>%
+  dplyr::filter(family != "Hydrobatidae (Northern Storm-Petrels)") %>%
[…]

### 2019-04-09 paper/main.tex
no message

@@
 \begin{abstract}
-Citizen science is important. yata yata yata
+Citizen science is so so important. yata yata yata
 \end{abstract}
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{T}
+\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0, allowing the tracking of progress, and assessment of whether biodiversity targets are being met.
+Unsurprisingly, trend detection is best estimated when using data from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value}.
+But scientific funding for long-term ecological research is rapidly disappearing [REFS???].
+Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.
 
-This work is really important \cite{theobald2015global}.
+Citizen science projects fall along a continuum of the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K.
+Butterfly Monitoring Scheme).
+All citizen science projects have associated biases, including spatial and temporal biases \cite{boakes2010distorted}, and different levels of observers' skills \cite{kelling2015can}.
+But these biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
+Regardless of biases, estimating trends is best done with structured projects \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, @kery2010site, @horns2018using, @van2013occupancy, @pagel2014quantifying}.
+
+The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential of these data in the future of ecology, conservation, and resource management \cite{pocock2018vision, mckinley2017citizen}.
+But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.
+
+Clearly, citizen science holds great potential for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
+But can it be improved?
+Are there mechanisms which can help diminish the spatial and temporal biases that exist in citizen science projects and associated data?
+
+Here, we investigate these questions, and our objective is to optimize citizen science sampling for trend detection of biodiversity.
+Many citizen science projects have specific biases and quality concerns.
+For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not all, citizen science projects.
+However, generalized across all citizen science projects are spatial and temporal biases \cite{boakes2010distorted}.
+For instance, biases such as intra-week differences in sampling \cite{courter2013weekend}, sampling near roads and high-density population centers \cite{kelling2015taking}, and XXXX are biases that are generalizable across citizen science projects.
+
+We predicted that the following parameters would be positively associated with the marginal value of a given citizen science observation: whether a site was sampled or not, median sampling interval, days since last sample, distance to the nearest sampled site, nearest neighbor sampling interval, and the number of unique days sampled.
+
+\begin{itemize}
+  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not.
+  \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site.
+  \item \textbf{Days since last sample}: the number of days between samples at a site.
+  \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site.
+  \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor.
+  \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site.
+\end{itemize}
+
+Here, we illustrate our vision using a popular citizen science project --- eBird \cite{sullivan2009ebird}.
 
 \section*{Results}
+
 \section*{Discussion}
+Citizen science projects need to continuously adapt to be successful, and to better inform conservation science.
+Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.
+
+The same approach is robust to spatial scale: whether a citizen science project is global, regional, or carried out in the constraints of a local park, the site can be gridded and observations can then be guided into the future.
+
+We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day.
+For every day, a citizen scientists can be presented with a map, showing the highest valued sites that should be visited.
+Imagine visitor centers at National Parks and
+
+Feedback to citizen science participants is important!
+Imagine visitor centers across the world at national parks or urban greenspaces providing users with a localized map showing which trail someone should visit if they are interested in contributing to citizen science.
+More broadly, this approach could be applied to anyone in the world.
+The global pull of ecotourism is increasing exponentially, allowing the the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled.
+
+This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations.
+
+Here, our framework focused on a specific statistical outcome: trend detection.
+Many other ecological outcomes arise from citizen science, including species distribution models, phylogeographical research, invasive species detection, or phenological research.
+These potential outcomes will have different optimal sampling strategies in space and time.
+We note that our framework is robust, with the necessary piece of information being a statistical model to extract leverage.
+
+This approach relies on statistical leverage, assigning every citizen science observation a measure of influence that influences a statistical model.
+
+- Looking forward but using the past
+- We only looked at trend detection, but many other models can be used in this process (e.g., species distribution models)
+- Other potentials for this general approach (identifying outliers in the dataset or misidentified species)
 
@@
 
-\subsection*{eBird data}
-We used eBird data....
+\subsection*{Study region} We illustrate our concepts throughout the Greater Sydney Region, and we gridded the region into 411 grids, all of which are ~ km.
+We used the R statistical environment to carry out all analyses.
 
-\subsection*{Parameter calculation}
-For 2018, we
+\subsection*{eBird data} We downloaded the eBird basic dataset (VERSION), and subsetted the data between January 1st, 2010 to December 31st, 2018.
+eBird collects data in the form of 'checklists' --- a list of all species identified (audibly or visually) for a given spatiotemporal coordinates.
+eBird relies on an extensive network of regional reviewers who are local experts of the avifauna [REF]
+
+\subsection*{Trend detection model} For illustration purposes, we fitted a simple generalized linear model, based on presence/absence for each of 274 species --- those of which had > 20 observations in the region throughout the study period.
+The GLM took the form of: XXXXXX.
+County was included to add a spatial representation throughout the Greater Sydney Region, while we included the length of the list in the model as an offset to account for the effort associated with each individual eBird checklist.
+
+We acknowledge that a number of different models can be fit which assesses trend detection [REFS], and biases can be accounted for in a variety of ways [REFS].
+These models were fitted for the period between 2010-2018.
+
+\subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable [REFS].
+In other words, it is a measure of how much an observation influences the results of an observation.
+In our instance, because we had multiple predictor variables in our GLMs, we used dfBeta as a measure of statistical leverage.
+dfBeta can be calculated for a given parameter in a regression model.
+
+For each species (N=274) we calculated the dfBeta for each of the 29,055 observations used for each model.
+Then for every checklist assessed in the model, we took the sum of the absolute value of the dfBetas for that checklist as a measure of a checklist's leverage in understanding species' trends throughout the Greater Sydney Region.
+
+\subsection*{Parameter calculation} For each day in 2018 (N=365), for each grid, we dynamically calculated the following parameters, based on sampling: (1) whether a grid cell was sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date.
+These are some hypothesized parameters which we felt would influence the relative value of a given sampling event.
+
+We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids.
+We ran a linear regression to parameterize what parameters had the greatest impact on the leverage of a given observation.
+
+\subsection*{Expected leverage calculation} We XXXXX.
+
 }

### 2019-04-09 paper/main.tex
bit more crap on paper

@@
 
-\title{Optimizing biodiversity sampling by citizen scientists}
+\title{Optimizing the future of biodiversity sampling by citizen scientists}
 
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0, allowing the tracking of progress, and assessment of whether biodiversity targets are being met.
-Unsurprisingly, trend detection is best estimated when using data from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value}.
-But scientific funding for long-term ecological research is rapidly disappearing [REFS???].
+\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0.
+Reliable biodiversity estimates allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}.
+Unsurprisingly, reliable trend detection is usually derived from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value, magurran2010long}.
+But scientific funding for long-term ecological and conservation research is rapidly disappearing [REFS???].
 Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.
 
-Citizen science projects fall along a continuum of the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K.
+Citizen science --- the cooperation between a range of experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}.
+Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K.
 Butterfly Monitoring Scheme).
-All citizen science projects have associated biases, including spatial and temporal biases \cite{boakes2010distorted}, and different levels of observers' skills \cite{kelling2015can}.
-But these biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
-Regardless of biases, estimating trends is best done with structured projects \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, @kery2010site, @horns2018using, @van2013occupancy, @pagel2014quantifying}.
+The level of structure, in turn, influences the biases and data quality concerns of a particular project.
+For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects.
+Generalizeable among citizen science projects, however, are various spatial and temporal biases \cite{boakes2010distorted, bird2014statistical}.
+Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity are all examples of spatial and temporal biases.
+Importantly, these biases are not restricted to citizen science projects.
+'Professional scientists' also have biased sampling of biodiversity, highlighted by biased museum collections [REFS???].
 
-The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential of these data in the future of ecology, conservation, and resource management \cite{pocock2018vision, mckinley2017citizen}.
+Fortunately, in the case of broad-scale biodiversity data collected by citizen scientists, these various biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
+Despite these biases, citizen science data have increased our knowledge on species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}.
+Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
+
+The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
 But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.
 
-Clearly, citizen science holds great potential for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
-But can it be improved?
-Are there mechanisms which can help diminish the spatial and temporal biases that exist in citizen science projects and associated data?
+However, these 'big-data' statistical approaches \cite{kelling2015taking} are most applicable for data originating from large, successful citizen science projects.
+What about projects that are just starting?
+Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}?
+Are there optimal strategies for sampling in space and time, which are best able to contribute to estimating biodiversity trends?
 
-Here, we investigate these questions, and our objective is to optimize citizen science sampling for trend detection of biodiversity.
-Many citizen science projects have specific biases and quality concerns.
-For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not all, citizen science projects.
-However, generalized across all citizen science projects are spatial and temporal biases \cite{boakes2010distorted}.
-For instance, biases such as intra-week differences in sampling \cite{courter2013weekend}, sampling near roads and high-density population centers \cite{kelling2015taking}, and XXXX are biases that are generalizable across citizen science projects.
-
-We predicted that the following parameters would be positively associated with the marginal value of a given citizen science observation: whether a site was sampled or not, median sampling interval, days since last sample, distance to the nearest sampled site, nearest neighbor sampling interval, and the number of unique days sampled.
+Here, we investigate these questions with a specific objective: investigate how spatial and temporal sampling influences trend detection of biodiversity.
+Our approach is site-specific and dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation for a particular day.
+Here, we summarize the parameters in space and time which we hyptohesized would influence the value of a citizen science observation.
 
 \begin{itemize}
-  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not.
+  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than sampled previously sampled sites.
   \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site.
+We predicted that the median sampling interval would be positively associated with the value of a citizen science observation.
+I.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
   \item \textbf{Days since last sample}: the number of days between samples at a site.
+We predicted that the days since the last sample would positively associate with the value of a citizen science observation.
   \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site.
+We hypothesized that for the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
   \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor.
+We hypothesized that this would be positively influence the value of a observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value observations.
   \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site.
+We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with lots of observations would receive value given the long-term data originating from them.
 \end{itemize}
 
-Here, we illustrate our vision using a popular citizen science project --- eBird \cite{sullivan2009ebird}.
+We illustrate our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and data from the Greater Sydney Region, NSW, Australia.
 
-\section*{Results}
+\section*{Results and Discussion}
+We found weak evidence that unsampled sites were more marginally valuable than unsampled sites across grid sizes (Fig.
+S1), noting, however, that there were no unsampled grids in the 50 km grid cell size.
+This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- [REF] may not be the most applicable approach for detecting trends using citizen science data.
+This also suggests that citizen scientists generally are sufficiently sampling the appropriate areas: local knowledge indicates where birders should go birding.
 
-\section*{Discussion}
+However, there are still parameters that are significantly important for the marginal value of a citizen science observation (PARAMETER ESTIMATES FIGURE).
+At all four grid sizes, days since the last sample was positively associated with the marginal value of a citizen science observation.
+For all but the 10 km grid size, distance to the nearest sampled grid was positively associated with the marginal value of a citizen science observation.
+Conversely, the results for median sampling interval were negatively associated with the value of a citizen science observation.
+Nearest neighbor sampling interval had relatively weak influence on the value of a citizen science observation, compared with other parameters, as did the number of unique days sampled for a given grid.
+
+These results suggest that the value of a given citizen science observation can indeed be predicted based on temporal and spatial sampling.
+Through time, the longer a grid goes unsampled (i.e., days since last sample), that grid's marginal value will continue to increase.
+Similarly, this is synergistic with the distance to the nearest sampled grid, whereby the further away from a sampled grid an observation is, the more valuable that observation is.
+
+We were also able to use our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day.
+This approach required us to look backward first, in order to predict future values.
+(STATIC MAP FIGURE).
+
 Citizen science projects need to continuously adapt to be successful, and to better inform conservation science.
@@
 
+\subsection*{Data availability}
+
 }

### 2019-04-10 paper/main.tex
More more more!

@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{Please provide details of author contributions here.}
-\authordeclaration{The authors have no conflict of interest to declare.}
+\authorcontributions{Author contributions: C.T.C., W.K.C. designed research, performed research, and analyzed research. C.T.C., W.K.C., J.J.L.R., A.G.B.P., and R.E.M. wrote the paper.}
+\authordeclaration{The authors declare no conflict of interest.}
 
@@
 
+We found generally consistent results, based on the grid size chosen, highlighting the influence of spatial scale of our approach.
+Ultimately, this approach could be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park; the site can be gridded and observations can then be guided into the future.
+It is likely that the spatial grid cells would need to change, relative to the spatial scale of the study.
+
 Citizen science projects need to continuously adapt to be successful, and to better inform conservation science.
 Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.
-
-The same approach is robust to spatial scale: whether a citizen science project is global, regional, or carried out in the constraints of a local park, the site can be gridded and observations can then be guided into the future.
-
 We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day.
-For every day, a citizen scientists can be presented with a map, showing the highest valued sites that should be visited.
-Imagine visitor centers at National Parks and
-
 Feedback to citizen science participants is important!
+This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations.
 Imagine visitor centers across the world at national parks or urban greenspaces providing users with a localized map showing which trail someone should visit if they are interested in contributing to citizen science.
-More broadly, this approach could be applied to anyone in the world.
+More broadly, this approach could be applied to anyone in the world, and citizen science participants of a given project could opt in to receive daily, weekly, or monthly updates on where they should sample.
 The global pull of ecotourism is increasing exponentially, allowing the the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled.
-
-This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations.
 
@@
 These potential outcomes will have different optimal sampling strategies in space and time.
+For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids than for species trend detection.
 We note that our framework is robust, with the necessary piece of information being a statistical model to extract leverage.
 
-This approach relies on statistical leverage, assigning every citizen science observation a measure of influence that influences a statistical model.
+Arguably, citizen science will continue to shape the future of ecology and conservation, as it has substantially for the past 100 years [REFS].
+But we need to look forward the future: how can we maximize the immense volunteer effort?
+In eBird alone, citizen scientists contributed XXXX number of volunteer hours focused on looking for birds.
+Ultimately, birders do a good job of sampling, but we believe that we can still optimize the collective citizen science effort.
+We provide a framework with which citizen science project managers can use fit-to-purpose.
 
-- Looking forward but using the past
-- We only looked at trend detection, but many other models can be used in this process (e.g., species distribution models)
-- Other potentials for this general approach (identifying outliers in the dataset or misidentified species)
+\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km^{2} (Fig.
+SX), where a grid represented a 'site'.
+We used the R statistical environment \cite{rcoreteam2018r} to carry out all analyses, relying heavily on the tidyverse \cite{wickham2017tidyverse}, ggplot2 \cite{wickham2016ggplot}, and sf \cite{pebesma2018sf} packages.
 
-\matmethods{Please describe your materials and methods here.
-This can be more than one paragraph, and may contain subsections and equations as required.
-Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.
+In order to test our predictions, we relied on the eBird basic dataset (version ebd_relDec-2018; available at: <https://ebird.org/data/download>), subsetting the data between January 1^{st}, 2010 to December 31^{st}, 2018.
+eBird is a particular successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
+eBird relies on volunteer birdwatchers who collect data in the form of 'checklists' --- a list of all species identified (audibly or visually) for given spatiotemporal coordinates.
+eBird relies on an extensive network of regional reviewers who are local experts of the avifauna \cite{gilfedder2019brokering} to ensure data quality \cite{sullivan2009ebird}.
 
-\subsection*{Study region} We illustrate our concepts throughout the Greater Sydney Region, and we gridded the region into 411 grids, all of which are ~ km.
-We used the R statistical environment to carry out all analyses.
+\subsection*{Trend detection model} We first filtered the eBird basic dataset \cite{callaghan2017assessing, johnston2018estimates, la2014role}, by the following criteria: (1) we only included complete checklists, (2) only included terrestrial bird species, (3) removed any nocturnal checklists, (4) only included checklists which were > 5 minutes and < 240 minutes in duration, (5) only included checklists which travelled < 5 km or covered < 500 Ha, and (6) only included checklists which had > 4 species on it, as checklists with less than 4 species were likely to be targeted searches for particular species \cite{walker2017using, szabo2010regional}.
 
-\subsection*{eBird data} We downloaded the eBird basic dataset (VERSION), and subsetted the data between January 1st, 2010 to December 31st, 2018.
-eBird collects data in the form of 'checklists' --- a list of all species identified (audibly or visually) for a given spatiotemporal coordinates.
-eBird relies on an extensive network of regional reviewers who are local experts of the avifauna [REF]
+For any species with > 50 observations (N=235), we fit a generalized linear model using the 'glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}.
+The models consisted of a continuous term for day, beginning on January 1^{st}, 2010, and a categorical term for county, providing a spatial component to the models.
+We also included an offset term for the number of species seen on a given eBird checklist, accounting for temporal and spatial effort of that checklist \cite{szaboe2010regional}.
+Each of the models were fit with a binomial family distribution.
+A total of 25,995 observations (i.e., eBird checklists) were used to fit each of the models.
 
-\subsection*{Trend detection model} For illustration purposes, we fitted a simple generalized linear model, based on presence/absence for each of 274 species --- those of which had > 20 observations in the region throughout the study period.
-The GLM took the form of: XXXXXX.
-County was included to add a spatial representation throughout the Greater Sydney Region, while we included the length of the list in the model as an offset to account for the effort associated with each individual eBird checklist.
+\subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable \cite{cook1977detection}.
+In other words, it is a measure of how much a given observation influences the results of a statistical outcome.
+In our instance, we had multiple predictor variables in our GLMs, and so we used dfBeta as a measure of statistical leverage for each observation.
+dfBeta measures the change to the observed parameters of a new regression equation, after omitting the \textit{i^{th}} observation from the dataset \cite{belsley1980regression}.
+It follows the formula:
 
-We acknowledge that a number of different models can be fit which assesses trend detection [REFS], and biases can be accounted for in a variety of ways [REFS].
-These models were fitted for the period between 2010-2018.
+\begin{equation}
+D F B E T A=\hat{\beta}-\beta_{(i)}=\frac{\left(X^{\prime} X\right)^{-1} X_{i} r_{i}}{1-h_{i}}
+\end{equation}
 
-\subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable [REFS].
-In other words, it is a measure of how much an observation influences the results of an observation.
-In our instance, because we had multiple predictor variables in our GLMs, we used dfBeta as a measure of statistical leverage.
-dfBeta can be calculated for a given parameter in a regression model.
+where X is th predictor variable matrix, \textit{r} the residual vector, \textit{i h} the \textit{i^{th}} diagonal member, and \textit{i x} the \textit{i^{th}} line of matrix X.
+The value of dfBeta proportionally decreases with increasing number of observations.
 
-For each species (N=274) we calculated the dfBeta for each of the 29,055 observations used for each model.
-Then for every checklist assessed in the model, we took the sum of the absolute value of the dfBetas for that checklist as a measure of a checklist's leverage in understanding species' trends throughout the Greater Sydney Region.
+In our case, each observation for a given species recieved a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the 'dfbetas' function from R \cite{rcoreteam2018r}.
+The measure of statistical leverage, then, of a given checklist was the sum of the absolute value of the dfBeta measures for each species (i.e., the sum of all 235 dfBetas).
+This measure of statistical leverage was thus a measure of a checklist's influence in understanding cumulative species' trends throughout the Greater Sydney Region.
 
-\subsection*{Parameter calculation} For each day in 2018 (N=365), for each grid, we dynamically calculated the following parameters, based on sampling: (1) whether a grid cell was sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date.
-These are some hypothesized parameters which we felt would influence the relative value of a given sampling event.
+\subsection*{Parameter calculation} After our model was fit from 2010---2018, we calculated the predicted parameters of interest for each day in 2018 (N=365).
+For each individual grid, at each of the grid sizes, we dynamically calculated the following parameters: (1) whether a grid cell had ever been sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date, and (7) the number of unique sampling days within the grid cell.
 
-We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids.
-We ran a linear regression to parameterize what parameters had the greatest impact on the leverage of a given observation.
+We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids on that day.
+We ran a linear regression, for each of the different grid sizes considered in the analysis, to investigate what parameters were of significant interest.
+Prior to modelling, duration was highly correlated with median sampling interval for the majority of the grid size analyses, and as such, was excluded from consdieration.
+Given the paramters' correlation varies among grid cell sizes, we need to ensure a robust, and simple model.
+All variables were standardized prior to modelling, ensuring that the effect sizes of the given parameters were meaningful.
+The response variable, dfBeta (i.e., value) was log-transformed prior to modelling to meet model assumptions.
+Thus, the final model included a log-tranformed dfBeta response variable, regressed against standardized median sampling interval, number of days sampled, days since the last sample, distance to the nearest sampled neighbor, and the neighbor's median sampling interval.
 
-\subsection*{Expected leverage calculation} We XXXXX.
+After our model was fit, we used the 'augment' function from the broom package \cite{robinson2018broom} to predict the expected leverage for every grid cell in the Greater Sydney Region, for every day.
+For grid cells which were unsampled, we assigned them the mean of the sampled grid cells, based on our lack of evidence that unsampled cells were significantly more valuable than sampled cells.
+This prediction was done for every day of 2018.
 
-\subsection*{Data availability}
-
+\subsection*{Data availability} eBird data are available for download, but the necessary portion of the dataset, along with spatial data, and code to reproduce our analyses are available at: GitHub repository.
 }

### 2019-04-10 paper/main.tex
getting somewhere with an overall framework and approach to the paper!

@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{Citizen scientists are increasingly monitoring biodiversity, but we need to better understand how to optimally sample biodiversity in time and space for intended outcomes, relying on citizen science data.}
+\significancestatement{Citizen scientists are increasingly relied on for monitoring of biodiversity, but we need to better understand how to optimally sample biodiversity in time and space for specific outcomes, relying on citizen science data.
+We provide a framework which optimizes citizen science sampling, specific to an intended statistical outcome.}
 
@@
 \begin{abstract}
-Citizen science is so so important. yata yata yata
+Monitoring biodiversity trends in space and time is critical for conservation assessment and threat mitigation.
+Increasingly, scientists and conservationists are turning to citizen science data to extract information about population trends.
+But, citizen science data are generally 'noisy' with a suite of biases, seen these biases are seen as a necessary hurdle when using these data.
+In order to continue to rely on citizen science in the future of biodiversity monitoring, we aimed to provide a generalized framework and tested how a suite of temporal and spatial sampling biases influence the marginal value of a given citizen science observation --- borrowing from the theory of statistical leverage.
+We find that for a given site, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation.
+We introduce a dynamic framework, whereby we can predict expected marginal value (i.e., expected leverage) for a given site in space and time.
+Ultimately, we envision an approach where citizen scientists are guided and incentivized to submit sampling at sites with the highest expected marginal value.
 \end{abstract}
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0.
-Reliable biodiversity estimates allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}.
-Unsurprisingly, reliable trend detection is usually derived from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value, magurran2010long}.
+\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering, yoccoz2001monitoring} \parshape=0.
+Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}.
+Unsurprisingly, reliable trend estimates are best derived from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value, magurran2010long}.
 But scientific funding for long-term ecological and conservation research is rapidly disappearing [REFS???].
@@
 Citizen science --- the cooperation between a range of experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}.
-Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K.
+Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects with little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K.
 Butterfly Monitoring Scheme).
@@
 Generalizeable among citizen science projects, however, are various spatial and temporal biases \cite{boakes2010distorted, bird2014statistical}.
-Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity are all examples of spatial and temporal biases.
-Importantly, these biases are not restricted to citizen science projects.
-'Professional scientists' also have biased sampling of biodiversity, highlighted by biased museum collections [REFS???].
+Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of spatial and temporal biases.
+These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased, evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}.
+Indeed, many sampling methods have been proposed to optimally sample biodiversity \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
+But little attention has been given to optimal sampling by citizen scientists \cite{harrison2014assessing}.
 
-Fortunately, in the case of broad-scale biodiversity data collected by citizen scientists, these various biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
-Despite these biases, citizen science data have increased our knowledge on species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}.
-Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
+This is probably, in part, because these spatial and temporal biases have been seen as a 'necessary evil', associated with citizen science data \cite{parrish2018exposing}.
+Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), these various biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
+Indeed, despite known biases, citizen science data have increased our knowledge of species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}.
+Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}.
+But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
+From a conservation perspective, the goal is relatively straightforward: provide robust measures of a species' trend through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}.
+The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of the trend estimates (e.g., Fig.
+2).
+This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and is generally proportional to the number of observations, and appropriate sampling, through time (e.g., Video S1).
 
-The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
+The number of citizen science projects which are focused on ecological and environmental monitoring is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
 But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.
 
-However, these 'big-data' statistical approaches \cite{kelling2015taking} are most applicable for data originating from large, successful citizen science projects.
+As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data.
 What about projects that are just starting?
 Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}?
-Are there optimal strategies for sampling in space and time, which are best able to contribute to estimating biodiversity trends?
+Are there optimal strategies for sampling in space and time for estimating biodiversity trends?
 
@@
 Our approach is site-specific and dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation for a particular day.
-Here, we summarize the parameters in space and time which we hyptohesized would influence the value of a citizen science observation.
+We test our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird biodiversity observations from the Greater Sydney Region, NSW, Australia.
+First, we summarize the parameters in space and time which we hyptohesized would influence the marginal value of a citizen science observation.
 
@@
 
-We illustrate our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and data from the Greater Sydney Region, NSW, Australia.
+\section*{Results and Discussion}
+We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled grid cell (Fig.
+S1).
+To assess the robustness of these results, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids.
+For the 5 km (p=0.669), and 10 km (p=0.093) grid sizes, there was an insigificant relationship, but for the 25 km (p=0.035) grid size, there was a statistically significant relationship.
+At the 50 km grid sizes, the test was not possible because all grids had been sampled.
+This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
+In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling the biodiversity in space.
 
-\section*{Results and Discussion}
-We found weak evidence that unsampled sites were more marginally valuable than unsampled sites across grid sizes (Fig.
-S1), noting, however, that there were no unsampled grids in the 50 km grid cell size.
-This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- [REF] may not be the most applicable approach for detecting trends using citizen science data.
-This also suggests that citizen scientists generally are sufficiently sampling the appropriate areas: local knowledge indicates where birders should go birding.
-
-However, there are still parameters that are significantly important for the marginal value of a citizen science observation (PARAMETER ESTIMATES FIGURE).
-At all four grid sizes, days since the last sample was positively associated with the marginal value of a citizen science observation.
-For all but the 10 km grid size, distance to the nearest sampled grid was positively associated with the marginal value of a citizen science observation.
-Conversely, the results for median sampling interval were negatively associated with the value of a citizen science observation.
-Nearest neighbor sampling interval had relatively weak influence on the value of a citizen science observation, compared with other parameters, as did the number of unique days sampled for a given grid.
-
-These results suggest that the value of a given citizen science observation can indeed be predicted based on temporal and spatial sampling.
+However, for those sampled sites, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
+2).
+For all four grid sizes, days since the last sample in a grid was positively associated with the marginal value of a sampling event, and this parameter had the strongest effect size for all but the 5 km grid size models.
+Similarly, for all but the 10 km grid size, distance to the nearest sampled grid was positively associated with the marginal value of a sampling event.
+Surprisingly, the results for median sampling interval were negatively associated with the marginal value of a sampling event.
+Nearest neighbor sampling interval and the total number of uniquely sampled days had relatively weak influence on the value of a sampling event, compared with other parameters (i.e., smaller effect sizes).
+Together, these results suggest that there are possible improvements in biodiversity sampling for trend detection.
 Through time, the longer a grid goes unsampled (i.e., days since last sample), that grid's marginal value will continue to increase.
 Similarly, this is synergistic with the distance to the nearest sampled grid, whereby the further away from a sampled grid an observation is, the more valuable that observation is.
+The effect of distance was strongest at the smallest grid size, suggesting that when more grids (i.e., sites) are included, that the distance to the nearest sample is increasingly important for monitoring biodiversity trends.
 
-We were also able to use our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day.
-This approach required us to look backward first, in order to predict future values.
-(STATIC MAP FIGURE).
+The fact that median samplin interval of a site, as well as the number of unique days sampled at a site were generally insignificant suggests that the 'history' of a given site (i.e., the number and frequency of observations at that site) does not influence the dynamic marginal value of a sampling event.
 
 We found generally consistent results, based on the grid size chosen, highlighting the influence of spatial scale of our approach.
-Ultimately, this approach could be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park; the site can be gridded and observations can then be guided into the future.
-It is likely that the spatial grid cells would need to change, relative to the spatial scale of the study.
+It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}.
+Our approach of relying on statistical leverage as a measure of the value of a citizen science observation can be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park.
+The site can be gridded --- the size of a grid, however, would be proportional to the spatial scale of the study --- and the framework implemented, updated dynamically.
 
-Citizen science projects need to continuously adapt to be successful, and to better inform conservation science.
-Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.
-We envision a dynamic approach in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day.
-Feedback to citizen science participants is important!
-This approach evolves citizen science projects, moving past leaderboards which show the number of species or number of records for a participant, and show the 'value' of a participant's observations.
-Imagine visitor centers across the world at national parks or urban greenspaces providing users with a localized map showing which trail someone should visit if they are interested in contributing to citizen science.
-More broadly, this approach could be applied to anyone in the world, and citizen science participants of a given project could opt in to receive daily, weekly, or monthly updates on where they should sample.
-The global pull of ecotourism is increasing exponentially, allowing the the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled.
+Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}.
+Generally, this feedback is in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
+We used our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
+3, Video S2).
+Unsurprisingly, the grids with the highest expected leverage correlated negatively with human population density (Fig.
+S3).
+This approach required us to look backward first, using a model with all observations for 2018, based on dfBetas calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day.
+We envision a dynamic approach (Video S2) in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
+In this instance, leaderboards would move past numbers of species or submission and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset.
+Instead of participants preferrentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
+For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to citizen science.
+The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, their observations can be maximized.
 
 Here, our framework focused on a specific statistical outcome: trend detection.
-Many other ecological outcomes arise from citizen science, including species distribution models, phylogeographical research, invasive species detection, or phenological research.
-These potential outcomes will have different optimal sampling strategies in space and time.
-For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids than for species trend detection.
-We note that our framework is robust, with the necessary piece of information being a statistical model to extract leverage.
+Many other ecological outcomes arise from citizen science, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}.
+These potential outcomes will have different optimal sampling strategies in space and time, but can still me quantified in the same framework we introduce here.
+For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids \cite{crawley2001scale} than for species trend detection.
+Our framework is robust, with the necessary piece of information being a statistical model derived based on an intended ecological/management outcome.
 
-Arguably, citizen science will continue to shape the future of ecology and conservation, as it has substantially for the past 100 years [REFS].
-But we need to look forward the future: how can we maximize the immense volunteer effort?
-In eBird alone, citizen scientists contributed XXXX number of volunteer hours focused on looking for birds.
-Ultimately, birders do a good job of sampling, but we believe that we can still optimize the collective citizen science effort.
-We provide a framework with which citizen science project managers can use fit-to-purpose.
+\subsection*{Conclusions}
+Since eBird inception in 2002, citizen science contributors have collectively contributed > 30 million effort hours.
+And this is only one citizen science project, demonstrating the cumulative effort put-forth by citizen scientists.
+Arguably, citizen science will continue to shape the future of ecology and conservation, as it has substantially for the past couple centuries \cite{silvertown2009new}, with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in the monitoring of biodiversity.
+But we need to look towards the future.
+Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future?
+We found that there are general rules which could help to guide citizen science participants to better sampling in space and time: the distance to the nearest sampled site and the days since the last sample at a site both positively correlate with the marginal value of a citizen science observation.
+Moreover, we demonstrate a framework which practicioners can implement to better optimize their sampling designs, specific to a citizen science project and an intended goal.
 
-\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km^{2} (Fig.
+\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$ (Fig.
 SX), where a grid represented a 'site'.
@@
 
-In order to test our predictions, we relied on the eBird basic dataset (version ebd_relDec-2018; available at: <https://ebird.org/data/download>), subsetting the data between January 1^{st}, 2010 to December 31^{st}, 2018.
+In order to test our predictions, we relied on the eBird basic dataset (version $ebd-relDec-2018$; available at: https://ebird.org/data/download), subsetting the data between January 1$^{st}$, 2010 to December 31$^{st}$, 2018.
 eBird is a particular successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
@@
 For any species with > 50 observations (N=235), we fit a generalized linear model using the 'glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}.
-The models consisted of a continuous term for day, beginning on January 1^{st}, 2010, and a categorical term for county, providing a spatial component to the models.
-We also included an offset term for the number of species seen on a given eBird checklist, accounting for temporal and spatial effort of that checklist \cite{szaboe2010regional}.
+The models consisted of a continuous term for day, beginning on January 1$^{st}$, 2010, and a categorical term for county, providing a spatial component to the models (e.g., Fig.
+1).
+We also included an offset term for the number of species seen on a given eBird checklist, accounting for temporal and spatial effort of that checklist \cite{szabo2010regional}.
 Each of the models were fit with a binomial family distribution.
@@
 In our instance, we had multiple predictor variables in our GLMs, and so we used dfBeta as a measure of statistical leverage for each observation.
-dfBeta measures the change to the observed parameters of a new regression equation, after omitting the \textit{i^{th}} observation from the dataset \cite{belsley1980regression}.
+dfBeta measures the change to the observed parameters of a new regression equation, after omitting the \textit{i$^{th}$} observation from the dataset \cite{belsley1980regression}.
 It follows the formula:
@@
 
-where X is th predictor variable matrix, \textit{r} the residual vector, \textit{i h} the \textit{i^{th}} diagonal member, and \textit{i x} the \textit{i^{th}} line of matrix X.
+where X is the predictor variable matrix, \textit{r} the residual vector, \textit{i h} the \textit{i$^{th}$} diagonal member, and \textit{i x} the \textit{i$^{th}$} line of matrix X.
 The value of dfBeta proportionally decreases with increasing number of observations.
@@
 
-\subsection*{Data availability} eBird data are available for download, but the necessary portion of the dataset, along with spatial data, and code to reproduce our analyses are available at: GitHub repository.
+\subsection*{Data availability} All eBird data are freely available for download (https://ebird.org/data/download), but the necessary portion of the eBird basic dataset, along with spatial data, and code to reproduce our analyses are available at: GitHub repository.
 }

### 2019-04-10 paper/main.tex
added a 'methods' para before results

@@
 Our approach is site-specific and dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation for a particular day.
-We test our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird biodiversity observations from the Greater Sydney Region, NSW, Australia.
 First, we summarize the parameters in space and time which we hyptohesized would influence the marginal value of a citizen science observation.
+
+\subsection*{predictions}
 
@@
 \end{itemize}
+
+We test our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird biodiversity observations from the Greater Sydney Region, NSW, Australia.
+We fit presence/absence population trend models to 235 species throughout this region, based on data from 2010 to 2018 (Fig.
+1), and calculate statistical leverage for every citizen science sampling event, for each species.
+The leverage is a measure of how 'valuable' a given observation is to the population trend model for each species.
+Each sampling event is then assigned a cumulative measure of leverage --- i.e., marginal value.
+We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend detection of biodiversity.
+Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid (i.e., site) in space and time, deriving a dynamic map of expected marginal value for every day in 2018.
 

### 2019-04-11 R/modelling.R
more modelling stuff.

@@ -25,20 +25,28 @@ analysis_data <- params %>%
     sampled=="no" ~ 0)
     )
 
+
 # make df for modelling
-hist(analysis_data$DATE_dfbetas, breaks=30)
-
-hist(log(analysis_data$DATE_dfbetas), breaks=30)
-
 # rescale variables
 analysis_data <- analysis_data %>%
+  mutate(dist_km_nn=replace(dist_km_nn, dist_km_nn==0, 0.01)) %>%
   group_by(grid_size) %>%
   mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
   mutate(s.duration=arm::rescale(duration)) %>%
-  mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
-  mutate(s.days_since=rescale(days_since)) %>%
-  mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
-  mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time))
+  mutate(s.Number_of_days_sampled=arm::rescale(Number_of_days_sampled)) %>%
+  mutate(s.days_since=arm::rescale(days_since)) %>%
+  mutate(s.dist_km_nn=arm::rescale(dist_km_nn)) %>%
+  mutate(s.neighbor_waiting_time=arm::rescale(log(neighbor_waiting_time)))%>%
+  mutate(l.s.median_waiting_time=arm::rescale(log(median_waiting_time))) %>%
+  mutate(l.s.duration=arm::rescale(log(duration))) %>%
+  mutate(l.s.Number_of_days_sampled=arm::rescale(log(Number_of_days_sampled))) %>%
+  mutate(l.s.days_since=arm::rescale(log(days_since))) %>%
+  mutate(l.s.dist_km_nn=arm::rescale(log(dist_km_nn))) %>%
+  mutate(l.s.neighbor_waiting_time=arm::rescale(log(neighbor_waiting_time)))
+
+hist(analysis_data$DATE_dfbetas, breaks=30)
+
+hist(log(analysis_data$DATE_dfbetas), breaks=30)
 
 # look at correlation among predictor variables
 cor_object5 <- analysis_data %>%
@@ -204,18 +212,18 @@ ggsave(filename="Figures/param_estimates.png", width=6, height=5, units="in")
 
 # explore visreg plots
 par(mfrow=c(2, 2))
[…]

### 2019-04-16 paper/main.tex
more polishing

@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{Citizen scientists are increasingly relied on for monitoring of biodiversity, but we need to better understand how to optimally sample biodiversity in time and space for specific outcomes, relying on citizen science data.
-We provide a framework which optimizes citizen science sampling, specific to an intended statistical outcome.}
+\significancestatement{Citizen science has the potential to significantly contribute to ecology, conservation, and natural resource management in the future.
+But we need to better understand how to maximize the value from the collective effort of Citizen scientists.
+We provide a framework, specific to an intended statistical outcome (e.g., population trend model), relying on statistical leverage to assign the marginal value of a given citizen science observation.
+This framework is dynamic, and can be calculated for a given day, providing a map of expected value for a given citizen science observation.
+We recommend citizen scientists, and citizen science project managers, adapt this framework to their needs, with a focus on maximizing the value of citizen science efforts for biodiversity management.}
 
@@
 \begin{abstract}
-Monitoring biodiversity trends in space and time is critical for conservation assessment and threat mitigation.
-Increasingly, scientists and conservationists are turning to citizen science data to extract information about population trends.
-But, citizen science data are generally 'noisy' with a suite of biases, seen these biases are seen as a necessary hurdle when using these data.
-In order to continue to rely on citizen science in the future of biodiversity monitoring, we aimed to provide a generalized framework and tested how a suite of temporal and spatial sampling biases influence the marginal value of a given citizen science observation --- borrowing from the theory of statistical leverage.
-We find that for a given site, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation.
-We introduce a dynamic framework, whereby we can predict expected marginal value (i.e., expected leverage) for a given site in space and time.
-Ultimately, we envision an approach where citizen scientists are guided and incentivized to submit sampling at sites with the highest expected marginal value.
+Monitoring biodiversity trends in space and time is critical for conservation, but funding for ecology and conservation research is disappearing.
+Scientists are increasingly turning to citizen science data to extract information about population trends, with the aims of providing management recommendations.
+But these data are generally 'noisy' with a suite of biases that need to be statistically accounted for while estimating population trends.
+Undoubtedly, the immense increase in citizen science projects, and data, will continue to form an important resource for monitoring biodiversity in the future.
+Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time.
+We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the end statistical goal: in this case, population trends.
+We test how a suite of dynamic characteristics quantifying sampling in space and time influence the marginal value of an observation.
+We find that for a given grid, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation.
+Our framework is also independent to spatial-scale, easily adapted within a National Park, or an entire state or country.
+We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018.
+Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value.
 \end{abstract}
@@
 Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}.
-Unsurprisingly, reliable trend estimates are best derived from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value, magurran2010long}.
-But scientific funding for long-term ecological and conservation research is rapidly disappearing [REFS???].
+Unsurprisingly, reliable trend estimates are best derived from long-term \cite{lindenmayer2012value, magurran2010long}, well-designed surveys, in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}.
+But scientific funding for long-term ecological and conservation research is failing to keep pace with conservation research efforts and needs \cite{bakker2010changing}.
 Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.
@@
 These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased, evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}.
-Indeed, many sampling methods have been proposed to optimally sample biodiversity \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
-But little attention has been given to optimal sampling by citizen scientists \cite{harrison2014assessing}.
+Indeed, many sampling methods have been proposed to optimally sample biodiversity by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
+Structured citizen science projects generally adapt some form of optimal sampling (e.g., stratified sampling), but little attention has been given to optimal sampling by citizen scientists in unstructured citizen science projects \cite{harrison2014assessing}.
 
@@
 The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of the trend estimates (e.g., Fig.
-2).
+1).
 This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and is generally proportional to the number of observations, and appropriate sampling, through time (e.g., Video S1).
+
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=.8\linewidth]{glm_examples_edited.png}
+\caption{The ultimate goal in understanding population trends is to decrease the standard error for a population trend model.
+Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right).}
+\label{fig1}
+\end{figure}
 
@@
 But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.
-
 As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data.
-What about projects that are just starting?
+What about projects that are just beginning?
 Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}?
@@
 
-\subsection*{predictions}
+\subsection*{Predictions}
 
@@
 We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend detection of biodiversity.
-Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid (i.e., site) in space and time, deriving a dynamic map of expected marginal value for every day in 2018.
+Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid in space and time, producing a dynamic map of expected marginal value for every day in 2018.
 
 \section*{Results and Discussion}
-We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled grid cell (Fig.
-S1).
+We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled grid cell.
 To assess the robustness of these results, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids.
@@
 This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
-In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling the biodiversity in space.
+In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling biodiversity in space.
+
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=.8\linewidth]{param_estimates.png}
+\caption{Parameter estimates for standardized linear models for each of the four grid sizes (5, 10, 25, and 50).}
+\label{fig2}
+\end{figure}
 
@@
 Similarly, this is synergistic with the distance to the nearest sampled grid, whereby the further away from a sampled grid an observation is, the more valuable that observation is.
-The effect of distance was strongest at the smallest grid size, suggesting that when more grids (i.e., sites) are included, that the distance to the nearest sample is increasingly important for monitoring biodiversity trends.
-
-The fact that median samplin interval of a site, as well as the number of unique days sampled at a site were generally insignificant suggests that the 'history' of a given site (i.e., the number and frequency of observations at that site) does not influence the dynamic marginal value of a sampling event.
+The effect of distance was strongest at the smallest grid size, suggesting that when more grids (i.e., sites) are included, the distance to the nearest sample is increasingly important for monitoring biodiversity trends.
+The fact that median sampling interval of a site, as well as the number of unique days sampled at a site were generally insignificant suggests that the 'history' of a given site (i.e., the number and frequency of observations at that site) does not influence the dynamic marginal value of a sampling event.
 
@@
 Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}.
-Generally, this feedback is in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
+This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
 We used our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
 3, Video S2).
-Unsurprisingly, the grids with the highest expected leverage correlated negatively with human population density (Fig.
-S3).
-This approach required us to look backward first, using a model with all observations for 2018, based on dfBetas calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day.
+Unsurprisingly, the grids with the highest expected leverage correlated negatively with human population density.
+This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day.
 We envision a dynamic approach (Video S2) in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
-In this instance, leaderboards would move past numbers of species or submission and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset.
+In this instance, leaderboards would move past numbers of species or submissions and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset.
 Instead of participants preferrentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
 For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to citizen science.
-The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, their observations can be maximized.
+The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.
+
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=.8\linewidth]{param_estimates.png}
+\caption{Six different days in 2018, showing a map of the Greater Sydney Region, and the predicted expected marginal value (i.e., statistical leverage) of biodiversity sampling that day.
+We predicted the expected value of citizen science observations for every day in 2018 (Video S2).}
+\label{fig3}
+\end{figure}
 
@@
 Many other ecological outcomes arise from citizen science, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}.
-These potential outcomes will have different optimal sampling strategies in space and time, but can still me quantified in the same framework we introduce here.
+These potential outcomes will have different optimal sampling strategies in space and time, but can still be quantified in the same framework we introduce here.
 For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids \cite{crawley2001scale} than for species trend detection.
-Our framework is robust, with the necessary piece of information being a statistical model derived based on an intended ecological/management outcome.
+Our framework is robust, with the necessary piece of information being a statistical model based on an intended ecological/management outcome.
 
 \subsection*{Conclusions}
-Since eBird inception in 2002, citizen science contributors have collectively contributed > 30 million effort hours.
-And this is only one citizen science project, demonstrating the cumulative effort put-forth by citizen scientists.
-Arguably, citizen science will continue to shape the future of ecology and conservation, as it has substantially for the past couple centuries \cite{silvertown2009new}, with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in the monitoring of biodiversity.
+Since eBird's inception in 2002, citizen science contributors have collectively contributed > 30 million effort hours.
+And this is only one citizen science project.
+Clearly, the cumulative effort put-forth by citizen scientists is immense and, arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple centuries \cite{silvertown2009new} --- with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in the monitoring of biodiversity.
 But we need to look towards the future.
 Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future?
-We found that there are general rules which could help to guide citizen science participants to better sampling in space and time: the distance to the nearest sampled site and the days since the last sample at a site both positively correlate with the marginal value of a citizen science observation.
+We highlighted general rules which could help to guide citizen science participants to better sampling in space and time: the distance to the nearest sampled site and the days since the last sample at a site both positively correlate with the marginal value of a citizen science observation.
 Moreover, we demonstrate a framework which practicioners can implement to better optimize their sampling designs, specific to a citizen science project and an intended goal.

### 2019-04-17 R/modelling.R
save model outputs for different models.

@@ -365,7 +365,11 @@ visreg(mod50.l, "l.s.median_waiting_time")
 
 
 
-
+# write out each model object
+saveRDS(mod5.l, file="Data/Modelling data/model_5.l_km_object.RDS")
+saveRDS(mod10.l, file="Data/Modelling data/model_10.l_km_object.RDS")
+saveRDS(mod25.l, file="Data/Modelling data/model_25.l_km_object.RDS")
+saveRDS(mod50.l, file="Data/Modelling data/model_50.l_km_object.RDS")
 
 
 

### 2019-04-17 R/predicting.R
predicting on the new models

@@ -3,7 +3,7 @@ library(tidyverse)
 library(arm)
 library(tidyr)
 
-load("Data/Modelling data/model_5_km_object.RData")
+mod_examp <- readRDS("Data/Modelling data/model_5.l_km_object.RDS")
 
 oneday <- readRDS("Data/Sampling dates 2018/5 km/2018-12-31.RDS")
 
@@ -16,24 +16,24 @@ add_dfbeta <- function(grid_size, date){
   # scale the df, according to how the
   # dataframe for modelling was scaled
   df2 <- df %>%
-    mutate(s.median_waiting_time=arm::rescale(median_waiting_time)) %>%
-    mutate(s.duration=arm::rescale(duration)) %>%
-    mutate(s.Number_of_days_sampled=rescale(Number_of_days_sampled)) %>%
-    mutate(s.days_since=rescale(days_since)) %>%
-    mutate(s.dist_km_nn=rescale(dist_km_nn)) %>%
-    mutate(s.neighbor_waiting_time=rescale(neighbor_waiting_time)) %>%
+    mutate(l.s.median_waiting_time=arm::rescale(log(median_waiting_time))) %>%
+    mutate(l.s.duration=arm::rescale(log(duration))) %>%
+    mutate(l.s.Number_of_days_sampled=arm::rescale(log(Number_of_days_sampled))) %>%
+    mutate(l.s.days_since=arm::rescale(log(days_since))) %>%
+    mutate(l.s.dist_km_nn=arm::rescale(log(dist_km_nn))) %>%
+    mutate(l.s.neighbor_waiting_time=arm::rescale(log(neighbor_waiting_time))) %>%
   # fill in the NA values for each of these variables
   # necessary for predicting. Fill in with mean of that column
     replace_na(list(
-      s.median_waiting_time=mean(.$s.median_waiting_time, na.rm=TRUE),
-      s.duration=mean(.$s.duration, na.rm=TRUE),
-      s.Number_of_days_sampled=mean(.$s.Number_of_days_sampled, na.rm=TRUE),
-      s.days_since=mean(.$s.days_since, na.rm=TRUE),
-      s.dist_km_nn=mean(.$s.dist_km_nn, na.rm=TRUE),
-      s.neighbor_waiting_time=mean(.$s.neighbor_waiting_time, na.rm=TRUE)
[…]

### 2019-04-17 R/plot_leverage.R
change scale and make new plots based on new data

@@ -58,7 +58,7 @@ ggplot()+
   theme(panel.grid.major=element_line(color="transparent"))+
   theme(axis.text=element_blank())+
   theme(axis.ticks=element_blank())+
-  scale_fill_gradientn(colours=pal, name="Value", trans="log")
+  scale_fill_gradientn(colours=plasma_pal, name="Value", trans="log")
   
 # potential figure for paper
 example_dates <- c("2018-01-01", "2018-03-01", "2018-05-01",
@@ -80,8 +80,8 @@ ggplot()+
   theme(axis.ticks=element_blank())+
   facet_wrap(~dynamic_date)+
   scale_fill_gradientn(colours=plasma_pal, name="Value", trans="log", 
-                       breaks=c(0.01, 0.05, 0.15, 0.5, 3, 25), labels=c(0.01, 0.05, 0.15, 0.5, 3, 25),
-                       limits=c(0.01, 30))
+                       breaks=c(0.004, 0.01, 0.02, 0.05, 0.15, 0.5), labels=c(0.004, 0.01, 0.02, 0.05, 0.15, 0.5),
+                       limits=c(0.003, 0.5))
 
 ggsave("Figures/example_maps.png", width=8, height=6, units="in")
 
@@ -106,8 +106,8 @@ for (i in dates_2018) {
     theme(axis.ticks=element_blank())+
     facet_wrap(~dynamic_date)+
     scale_fill_gradientn(colours=plasma_pal, name="Value", trans="log", 
-                         breaks=c(0.01, 0.05, 0.15, 0.5, 3, 25), labels=c(0.01, 0.05, 0.15, 0.5, 3, 25),
-                         limits=c(0.01, 30))
+                         breaks=c(0.003, 0.01, 0.02, 0.05, 0.15, 0.5), labels=c(0.003, 0.01, 0.02, 0.05, 0.15, 0.5),
+                         limits=c(0.003, 0.5))
   
   ggsave(plot, filename=paste0("C:/Users/CTC/Desktop/map_gif2/", i, ".png"), dpi=150)
 }

### 2019-04-17 R/modelling.R
new param estimated figure

@@ -208,7 +208,7 @@ bind_rows(estimates_5, estimates_10, estimates_25, estimates_50) %>%
   theme(panel.grid.major.y=element_blank())
 
 
-ggsave(filename="Figures/param_estimates.png", width=6, height=5, units="in")
+
 
 # explore visreg plots
 par(mfrow=c(2, 2))
@@ -350,6 +350,8 @@ bind_rows(estimates_5.l, estimates_10.l, estimates_25.l, estimates_50.l) %>%
   theme(axis.ticks=element_line(color="black"))+
   theme(panel.grid.major.y=element_blank())
 
+
+ggsave(filename="Figures/param_estimates.png", width=6, height=5, units="in")
 
 par(mfrow=c(2, 2))
 visreg(mod5.l, "l.s.days_since")

### 2019-04-17 R/species_pop_trend_example.R
update Noisy Miner gif

@@ -399,7 +399,7 @@ for (i in unique(as.character(as.Date(date_sequence2$OBSERVATION_DATE)))) {
     theme(axis.ticks=element_line(color="black"))+
     theme(panel.grid.major.y=element_blank())
   
-  ggsave(filename = paste0("C:/Users/CTC/Desktop/gif_species/", i, ".png"))
+  ggsave(filename = paste0("C:/Users/CTC/Desktop/gif_species/", i, ".png"), height=4.8, width=6.2, units="in", dpi=150)
   
 }
 

### 2019-04-17 paper/main.tex
minor changes
updated figures

@@
 \begin{abstract}
-Monitoring biodiversity trends in space and time is critical for conservation, but funding for ecology and conservation research is disappearing.
-Scientists are increasingly turning to citizen science data to extract information about population trends, with the aims of providing management recommendations.
+We are currently in the midst of the sixth extinction, and conservation biology and ecology are increasingly interested in tracking biodiversity trends in space and time.
+Yet, funding for ecology and conservation research is simultaneously disappearing.
+In turn, scientists are increasingly relying on citizen science data to extract information about population trends, with the aims of providing management recommendations.
 But these data are generally 'noisy' with a suite of biases that need to be statistically accounted for while estimating population trends.
@@
 \centering
-\includegraphics[width=.8\linewidth]{param_estimates.png}
+\includegraphics[width=.8\linewidth]{example_maps.png}
 \caption{Six different days in 2018, showing a map of the Greater Sydney Region, and the predicted expected marginal value (i.e., statistical leverage) of biodiversity sampling that day.

### 2019-04-18 paper/main.tex
first draft complete

--- 2019-04-16 paper/main.tex
@@
 \begin{abstract}
-Monitoring biodiversity trends in space and time is critical for conservation, but funding for ecology and conservation research is disappearing.
-Scientists are increasingly turning to citizen science data to extract information about population trends, with the aims of providing management recommendations.
-But these data are generally 'noisy' with a suite of biases that need to be statistically accounted for while estimating population trends.
-Undoubtedly, the immense increase in citizen science projects, and data, will continue to form an important resource for monitoring biodiversity in the future.
-Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time.
-We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the end statistical goal: in this case, population trends.
-We test how a suite of dynamic characteristics quantifying sampling in space and time influence the marginal value of an observation.
-We find that for a given grid, the days since the last sample and the distance from the nearest sampled grid are positively associated with the marginal value of a citizen science observation.
-Our framework is also independent to spatial-scale, easily adapted within a National Park, or an entire state or country.
+We are currently in the midst of the sixth great extinction, and understanding biodiversity trends in space and time is essential for curbing biodiversity loss and prioritizing limited resources for conservation.
+Concomitantly, funding for ecology and conservation research is disappearing.
+In turn, scientists are increasingly using citizen science data to extract information about population trends, with the aim of providing management recommendations.
+But these data are generally 'noisy' with a suite of redundancies and gaps arising from the associated structure of the project.
+Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time by citizen scientists.
+We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to overall population trends.
+We test how a suite of dynamic characteristics of sampling in space and time influence the marginal value of citizen science sampling events.
+We find that the `history' of a grid influences the future optimal sampling: the marginal value of a sampling event was most associated with the number of unique sampling days of a grid.
+We also find that grids should aim to decrease their median sampling interval.
+Our framework is robust to spatial-scale, easily adapted within a national park, entire state, or country.
 We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018.
@@
 Butterfly Monitoring Scheme).
-The level of structure, in turn, influences the biases and data quality concerns of a particular project.
+The level of structure, in turn, influences the redundancies and gaps in the data, as well as the data quality of a particular project.
 For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects.
-Generalizeable among citizen science projects, however, are various spatial and temporal biases \cite{boakes2010distorted, bird2014statistical}.
-Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of spatial and temporal biases.
-These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased, evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}.
+Generalizeable among citizen science projects, however, are various redundancies and gaps (i.e., spatial and temporal biases) \cite{boakes2010distorted, bird2014statistical}.
+Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of such data constraints.
+These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}.
 Indeed, many sampling methods have been proposed to optimally sample biodiversity by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
@@
 
-This is probably, in part, because these spatial and temporal biases have been seen as a 'necessary evil', associated with citizen science data \cite{parrish2018exposing}.
+This is probably, in part, because these redundancies and gaps in the data are seen as a 'necessary hurdle', associated with such data \cite{parrish2018exposing}.
 Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), these various biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
@@
 But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
-From a conservation perspective, the goal is relatively straightforward: provide robust measures of a species' trend through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}.
-The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of the trend estimates (e.g., Fig.
+From a conservation perspective, the goal is relatively straightforward: provide robust measures of species' trends through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}.
+The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of trend estimates (e.g., Fig.
 1).
-This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and is generally proportional to the number of observations, and appropriate sampling, through time (e.g., Video S1).
+This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and the standard error is generally proportional to the number of observations, and appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video S1}).
 
@@
 \includegraphics[width=.8\linewidth]{glm_examples_edited.png}
-\caption{The ultimate goal in understanding population trends is to decrease the standard error for a population trend model.
+\caption{The ultimate goal in understanding population trends is to minimize the standard error for a population trend model, providing more robust measures of population trends.
 Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right).}
@@
 
-Here, we investigate these questions with a specific objective: investigate how spatial and temporal sampling influences trend detection of biodiversity.
-Our approach is site-specific and dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation for a particular day.
-First, we summarize the parameters in space and time which we hyptohesized would influence the marginal value of a citizen science observation.
+Here, we investigate these questions with a specific objective: assess how spatial and temporal sampling influences trend detection of biodiversity.
+Our approach is dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation in both time and space.
+First, we summarize the parameters which we hyptohesized would influence the marginal value of a citizen science observation.
 
@@
 
-We test our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird biodiversity observations from the Greater Sydney Region, NSW, Australia.
+We test our predictions using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and > 5 million bird observations from the Greater Sydney Region, NSW, Australia.
 We fit presence/absence population trend models to 235 species throughout this region, based on data from 2010 to 2018 (Fig.
@@
 We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend detection of biodiversity.
-Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid in space and time, producing a dynamic map of expected marginal value for every day in 2018.
+Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.
 
 \section*{Results and Discussion}
-We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled grid cell.
-To assess the robustness of these results, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids.
-For the 5 km (p=0.669), and 10 km (p=0.093) grid sizes, there was an insigificant relationship, but for the 25 km (p=0.035) grid size, there was a statistically significant relationship.
-At the 50 km grid sizes, the test was not possible because all grids had been sampled.
-This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
+To assess the robustness of our approach, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids.
+We found weak evidence that visiting an unsampled grid was marginally more valuable than visiting an already sampled grid.
+There was no statistical significance for the 5 km (p=0.669), and 10 km (p=0.093) grid sizes, but there was for the 25 km (p=0.035) grid size.
+At the 50 km grid size, this test was not possible because all grids had been sampled.
+These results suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
 In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling biodiversity in space.
@@
 \includegraphics[width=.8\linewidth]{param_estimates.png}
-\caption{Parameter estimates for standardized linear models for each of the four grid sizes (5, 10, 25, and 50).}
+\caption{Parameter estimates for standardized linear models for each of the four grid sizes (5, 10, 25, and 50).
+Variables were log-transformed and then standardized, allowing for the direct comparison among effect sizes.}
 \label{fig2}
@@
 
-However, for those sampled sites, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
+For those sampled grids, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
 2).
-For all four grid sizes, days since the last sample in a grid was positively associated with the marginal value of a sampling event, and this parameter had the strongest effect size for all but the 5 km grid size models.
-Similarly, for all but the 10 km grid size, distance to the nearest sampled grid was positively associated with the marginal value of a sampling event.
-Surprisingly, the results for median sampling interval were negatively associated with the marginal value of a sampling event.
-Nearest neighbor sampling interval and the total number of uniquely sampled days had relatively weak influence on the value of a sampling event, compared with other parameters (i.e., smaller effect sizes).
+In general, our parameters were positively associated with the marginal value of a sampling event.
+Of our parameters of interest, the number of unique days sampled (i.e., the `history' of a grid) had the strongest, positive, effect size, and this was robust to the grid size comparisons.
+The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km grid size.
+Distance to the nearest sampled grid and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters.
+Surprisingly, days since the last sample had less influence than the other parameters, albeit a positive association.
+The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, likely highlights the value of `revisiting' a site in order to extract the maximum amount of information.
+
 Together, these results suggest that there are possible improvements in biodiversity sampling for trend detection.
-Through time, the longer a grid goes unsampled (i.e., days since last sample), that grid's marginal value will continue to increase.
-Similarly, this is synergistic with the distance to the nearest sampled grid, whereby the further away from a sampled grid an observation is, the more valuable that observation is.
-The effect of distance was strongest at the smallest grid size, suggesting that when more grids (i.e., sites) are included, the distance to the nearest sample is increasingly important for monitoring biodiversity trends.
-The fact that median sampling interval of a site, as well as the number of unique days sampled at a site were generally insignificant suggests that the 'history' of a given site (i.e., the number and frequency of observations at that site) does not influence the dynamic marginal value of a sampling event.
+In short, the `history' of a grid is important for considering whether to sample that grid: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable.
+Further, a secondary goal should be to decrease (i.e., left-shift the distribution) of the median sampling interval of the grids.
+The effect of number of days sampled was strongest at the largest grid size, suggesting that given all grids are sampled with some regularity --- as they are with larger grid sizes --- the importance of the `history' of a site is inflated.
 
-We found generally consistent results, based on the grid size chosen, highlighting the influence of spatial scale of our approach.
+We found generally consistent results, based on the grid size chosen, highlighting the robustness of spatial scale in our approach.
 It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}.
 Our approach of relying on statistical leverage as a measure of the value of a citizen science observation can be used regardless of whether a citizen science project is global, regional, or carried out in the constraints of a local park.
-The site can be gridded --- the size of a grid, however, would be proportional to the spatial scale of the study --- and the framework implemented, updated dynamically.
+In any instance, the area can be gridded --- the size of a grid, however, would be proportional to the spatial scale of the study --- and the framework implemented, updated dynamically.
 
@@
 We used our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
-3, Video S2).
-Unsurprisingly, the grids with the highest expected leverage correlated negatively with human population density.
+3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}).
 This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day.
 We envision a dynamic approach (Video S2) in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
-In this instance, leaderboards would move past numbers of species or submissions and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset.
+In this instance, leaderboards would move past numbers of species or submissions and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
 Instead of participants preferrentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
-For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to citizen science.
+For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science.
 The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.
@@
 \centering
-\includegraphics[width=.8\linewidth]{param_estimates.png}
-\caption{Six different days in 2018, showing a map of the Greater Sydney Region, and the predicted expected marginal value (i.e., statistical leverage) of biodiversity sampling that day.
-We predicted the expected value of citizen science observations for every day in 2018 (Video S2).}
+\includegraphics[width=.8\linewidth]{example_maps.png}
+\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued grid cells that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region.
+This is a dynamic approach, where the predictions are updated as new observations are submitted to the citizen science database, and we have done this for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}).}
 \label{fig3}
@@
 
-Here, our framework focused on a specific statistical outcome: trend detection.
-Many other ecological outcomes arise from citizen science, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}.
-These potential outcomes will have different optimal sampling strategies in space and time, but can still be quantified in the same framework we introduce here.
+We focused our framework on a specific statistical outcome: trend detection.
+Many other ecological outcomes arise from citizen science datasets, including species distribution models \cite{bradsworth2017species, van2013opportunistic}, phylogeographical research \cite{bahls2014new, drury2019continent}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, or phenological research \cite{la2014role, supp2015citizen}.
+Each potential outcome will have different optimal sampling strategies in space and time, likely with nuanced trade-offs between outcomes.
+But these different outcomes can still be quantified in the same framework we introduce here.
 For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids \cite{crawley2001scale} than for species trend detection.
@@
 \subsection*{Conclusions}
-Since eBird's inception in 2002, citizen science contributors have collectively contributed > 30 million effort hours.
+Since eBird's inception in 2002, citizen scientists have collectively contributed > 30 million effort hours.
 And this is only one citizen science project.
-Clearly, the cumulative effort put-forth by citizen scientists is immense and, arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple centuries \cite{silvertown2009new} --- with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in the monitoring of biodiversity.
+Clearly, the cumulative effort put-forth by citizen scientists is immense, and arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple centuries \cite{silvertown2009new} --- with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in monitoring of biodiversity.
 But we need to look towards the future.
 Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future?
-We highlighted general rules which could help to guide citizen science participants to better sampling in space and time: the distance to the nearest sampled site and the days since the last sample at a site both positively correlate with the marginal value of a citizen science observation.
+We highlighted general rules which could help to guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation.
 Moreover, we demonstrate a framework which practicioners can implement to better optimize their sampling designs, specific to a citizen science project and an intended goal.
 
-\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$ (Fig.
-SX), where a grid represented a 'site'.
+\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$, where a grid represented a 'site'.
 We used the R statistical environment \cite{rcoreteam2018r} to carry out all analyses, relying heavily on the tidyverse \cite{wickham2017tidyverse}, ggplot2 \cite{wickham2016ggplot}, and sf \cite{pebesma2018sf} packages.
 
-In order to test our predictions, we relied on the eBird basic dataset (version $ebd-relDec-2018$; available at: https://ebird.org/data/download), subsetting the data between January 1$^{st}$, 2010 to December 31$^{st}$, 2018.
-eBird is a particular successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
+In order to test our predictions, we used the eBird basic dataset (version $ebd-relDec-2018$; available at: https://ebird.org/data/download), subsetting the data between January 1$^{st}$, 2010 and December 31$^{st}$, 2018.
+eBird is a particularly successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
 eBird relies on volunteer birdwatchers who collect data in the form of 'checklists' --- a list of all species identified (audibly or visually) for given spatiotemporal coordinates.
@@
 For any species with > 50 observations (N=235), we fit a generalized linear model using the 'glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}.
-The models consisted of a continuous term for day, beginning on January 1$^{st}$, 2010, and a categorical term for county, providing a spatial component to the models (e.g., Fig.
+The models consisted of a continuous term for day, beginning January 1$^{st}$, 2010, and a categorical term for county, providing a spatial component to the models (e.g., Fig.
 1).
@@
 The measure of statistical leverage, then, of a given checklist was the sum of the absolute value of the dfBeta measures for each species (i.e., the sum of all 235 dfBetas).
-This measure of statistical leverage was thus a measure of a checklist's influence in understanding cumulative species' trends throughout the Greater Sydney Region.
+This measure of statistical leverage was thus a measure of a checklist's influence in understanding cumulative species' trends throughout the Greater Sydney Region, and accordingly represented the marginal value of that particular checklist.
 
@@
 Prior to modelling, duration was highly correlated with median sampling interval for the majority of the grid size analyses, and as such, was excluded from consdieration.
-Given the paramters' correlation varies among grid cell sizes, we need to ensure a robust, and simple model.
-All variables were standardized prior to modelling, ensuring that the effect sizes of the given parameters were meaningful.
+Given the paramters' correlation varied among grid cell sizes, we needed to ensure a robust, and simple model.
+All variables were log-transformed and then standardized prior to modelling, ensuring that the effect sizes of the given parameters were meaningful.
 The response variable, dfBeta (i.e., value) was log-transformed prior to modelling to meet model assumptions.
-Thus, the final model included a log-tranformed dfBeta response variable, regressed against standardized median sampling interval, number of days sampled, days since the last sample, distance to the nearest sampled neighbor, and the neighbor's median sampling interval.
+Thus, the final model included a log-tranformed dfBeta response variable, regressed against log-transformed standardized median sampling interval, number of days sampled, days since the last sample, distance to the nearest sampled neighbor, and the neighbor's median sampling interval.
 
@@
 For grid cells which were unsampled, we assigned them the mean of the sampled grid cells, based on our lack of evidence that unsampled cells were significantly more valuable than sampled cells.
-This prediction was done for every day of 2018.
+Where one grid had multiple predicted leverages (i.e., where a grid had more than one checklist in a day), we randomly sampled to one of these expected leverages.
+This prediction process was repeated for every day of 2018.
 
-\subsection*{Data availability} All eBird data are freely available for download (https://ebird.org/data/download), but the necessary portion of the eBird basic dataset, along with spatial data, and code to reproduce our analyses are available at: GitHub repository.
+\subsection*{Data availability} All eBird data are freely available for download (https://ebird.org/data/download), but the necessary portion of the eBird basic dataset, along with spatial data, and code to reproduce our analyses are available at: https://github.com/coreytcallaghan/optimize_citizen_science_obs.
 }

### 2019-04-25 paper/main.tex
changes from Jodi, RM, and some from Will

@@
 
-\title{Optimizing the future of biodiversity sampling by citizen scientists}
+\title{Optimizing future biodiversity sampling by citizen scientists}
 
@@
 \author[b,a]{Jodi J. L. Rowley}
-\author[a,c]{William K. Cornwell}
+\author[c,a]{William K. Cornwell}
 
 \affil[a]{Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}
-\affil[b]{Australian Museum Research Institute, Australian Museum, Sydney, 2000, NSW, Australia}
+\affil[b]{Australian Museum Research Institute, Australian Museum, Sydney, 2010, NSW, Australia}
 \affil[c]{Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, 2052, NSW, Australia}
@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{Citizen science has the potential to significantly contribute to ecology, conservation, and natural resource management in the future.
-But we need to better understand how to maximize the value from the collective effort of Citizen scientists.
-We provide a framework, specific to an intended statistical outcome (e.g., population trend model), relying on statistical leverage to assign the marginal value of a given citizen science observation.
-This framework is dynamic, and can be calculated for a given day, providing a map of expected value for a given citizen science observation.
-We recommend citizen scientists, and citizen science project managers, adapt this framework to their needs, with a focus on maximizing the value of citizen science efforts for biodiversity management.}
+\significancestatement{The challenge of monitoring biodiversity during the ongoing extinction crisis is unlikely to be matched by sufficient funding to professional scientists.
+Citizen science may fill much of the shortfall, but currently, citizen science sampling is haphazard, and we can do a better job maximizing both the marginal value of each person's effort and the total information content of a given dataset.
+We provide a dynamic framework to do this, showing where and when to sample biodiversity to maximize value per effort.
+We recommend that citizen science projects adapt this framework to their needs, with a focus on maximizing the marginal value of each citizen science contribution to inform biodiversity conservation goals.}
 
@@
 \begin{abstract}
-We are currently in the midst of the sixth great extinction, and understanding biodiversity trends in space and time is essential for curbing biodiversity loss and prioritizing limited resources for conservation.
-Concomitantly, funding for ecology and conservation research is disappearing.
-In turn, scientists are increasingly using citizen science data to extract information about population trends, with the aim of providing management recommendations.
-But these data are generally 'noisy' with a suite of redundancies and gaps arising from the associated structure of the project.
-Here, we ask whether these data can be improved --- maximizing the information content extracted from citizen science datasets --- by optimizing sampling in space and time by citizen scientists.
-We provide a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to overall population trends.
-We test how a suite of dynamic characteristics of sampling in space and time influence the marginal value of citizen science sampling events.
-We find that the `history' of a grid influences the future optimal sampling: the marginal value of a sampling event was most associated with the number of unique sampling days of a grid.
-We also find that grids should aim to decrease their median sampling interval.
-Our framework is robust to spatial-scale, easily adapted within a national park, entire state, or country.
-We conclude by using our parameterized model to produce dynamic maps, in space and time, predicting the expected marginal value in space for every day in 2018.
-Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value.
+We are currently in the midst of earth's sixth extinction event, and measuring biodiversity trends in space and time is essential for prioritizing limited resources for conservation and curbing biodiversity loss.
+At the same time, the scope of the necessary monitoring is massively overwhelming the funding for professional scientific monitoring.
+In response, scientists are increasingly using citizen science data to estimate population trends, with the aim of providing prioritizing conservation efforts.
+But citizen science data are 'noisy', with redundancies and gaps arising from unstructured human behaviors in space and time.
+Here, we ask whether the information in these data can be improved --- maximizing the information content extracted from citizen science datasets for trend estimation.
+We develop and execute a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to our understanding of overall population trends for a given geographic region.
+We then make this framework predictive, estimating the expected marginal value of future biodiversity observations.
+We find that past observations are useful in forecasting where high value observations will occur in the future.
+Interestingly, we find high marginal value in both `hotspots` which are frequently sampled locations and `coldspots` areas far from recent sampling, suggesting that an optimal sampling regime balances `hotspot` sampling with a spread across the landscape.
+Our framework is applicable at any spatial scale, easily adapted within a national park, entire country, or global.
+Because expected value is dependent on the behavior of all other citizen scientists, our model produces dynamic maps of expected marginal value through time.
+Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value, collectively maximizing our understanding of biodiversity in space and time, and our ability to conserve it.
 \end{abstract}
@@
 \dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering, yoccoz2001monitoring} \parshape=0.
-Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}.
-Unsurprisingly, reliable trend estimates are best derived from long-term \cite{lindenmayer2012value, magurran2010long}, well-designed surveys, in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}.
-But scientific funding for long-term ecological and conservation research is failing to keep pace with conservation research efforts and needs \cite{bakker2010changing}.
+Reliable biodiversity trend estimates, at multiple spatial scales \cite{soberon2007assessing}, allow us to track our global progress in curbing biodiversity loss while managing our scarce conservation resources \cite{harrison2014assessing}.
+Unsurprisingly, reliable trend estimates are best derived from long-term \cite{lindenmayer2012value, magurran2010long}, well-designed surveys, carried out over a wide spatial and temporal scale \cite{harrison2014assessing, vellend2017estimates, kery2009trend}.
+But scientific funding for long-term ecological and conservation research is failing to keep pace with conservation needs \cite{bakker2010changing, rios2018fieldwork}.
 Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.
 
-Citizen science --- the cooperation between a range of experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}.
-Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects with little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K.
+Citizen science --- the cooperation between scientific experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}.
+Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects which require little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K.
 Butterfly Monitoring Scheme).
-The level of structure, in turn, influences the redundancies and gaps in the data, as well as the data quality of a particular project.
-For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects.
+The level of structure, in turn, influences the degree of redundancies and gaps in the data, as well as the overall data quality of a particular project.
+For instance, observer skill \cite{kelling2015can}, number of participants in a group, and the technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects.
 Generalizeable among citizen science projects, however, are various redundancies and gaps (i.e., spatial and temporal biases) \cite{boakes2010distorted, bird2014statistical}.
-Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of such data constraints.
-These biases are not neccesarily restricted to citizen science projects, as our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}.
-Indeed, many sampling methods have been proposed to optimally sample biodiversity by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
-Structured citizen science projects generally adapt some form of optimal sampling (e.g., stratified sampling), but little attention has been given to optimal sampling by citizen scientists in unstructured citizen science projects \cite{harrison2014assessing}.
+Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of how unstructured human behavior leads to redundancies and gaps in citizen science data.
+These biases are not restricted to citizen science projects.
+Indeed, our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted} and many sampling methods have been proposed to optimize biodiversity sampling by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
+While structured citizen science projects often adapt some aspects of optimal sampling in their methods (e.g., stratified sampling), little attention has been given to optimal sampling in unstructured citizen science projects \cite{harrison2014assessing}.
 
-This is probably, in part, because these redundancies and gaps in the data are seen as a 'necessary hurdle', associated with such data \cite{parrish2018exposing}.
-Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), these various biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
+One of the reasons that optimal sampling has been largely ignored in unstructured citizen science projects is because redundancies and gaps in the data are seen as a "necessary hurdle" \cite{parrish2018exposing}.
+Also, in the case of broad-scale biodiversity data collected at large voluminous scales (e.g., eBird), the biases can generally be accounted for statistically \cite{isaac2014statistics, robinson2018correcting}, for instance, by filtering or subsetting data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}.
 Indeed, despite known biases, citizen science data have increased our knowledge of species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}.
-Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}.
+
+Despite the potential ecological outcomes from citizen science data, estimating biodiversity trends is perhaps the most important, given the current need to efficiently and effectively monitor biodiversity \cite{harrison2014assessing, honrado2016fostering, yoccoz2001monitoring}.
+From a conservation perspective, the goal is relatively straightforward: provide robust measures of species' trends through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}.
+Estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for, generally resulting in less standard errors) \cite{fox2011new}.
 But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
-From a conservation perspective, the goal is relatively straightforward: provide robust measures of species' trends through time, a critical component of the IUCN Red List Index \cite{baillie2008toward}.
-The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error of trend estimates (e.g., Fig.
+The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error surrounding these estimates (e.g., Fig.
 1).
-This is possible with citizen science data \cite{kery2010site, horns2018using, van2013occupancy, pagel2014quantifying} and the standard error is generally proportional to the number of observations, and appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video S1}).
+Unsurprisingly, the standard error is generally proportional to the number of observations, as well as appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video 1}).
 
@@
 \caption{The ultimate goal in understanding population trends is to minimize the standard error for a population trend model, providing more robust measures of population trends.
-Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right).}
+Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right), in the Greater Sydney Region, Australia.}
 \label{fig1}
@@
 As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data.
-What about projects that are just beginning?
+And even with 'big data', this is not always possible.
+But what about projects that are just beginning?
 Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}?
@@
 Our approach is dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation in both time and space.
-First, we summarize the parameters which we hyptohesized would influence the marginal value of a citizen science observation.
+First, we summarize the parameters which we hypothesized would influence the marginal value of a citizen science observation.
 
@@
 \begin{itemize}
-  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than sampled previously sampled sites.
+  \item \textbf{Site sampling}: if a site has been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than previously sampled sites.
   \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site.
-We predicted that the median sampling interval would be positively associated with the value of a citizen science observation.
-I.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
+We predicted that the median sampling interval would be positively associated with the value of a citizen science observation; i.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
   \item \textbf{Days since last sample}: the number of days between samples at a site.
-We predicted that the days since the last sample would positively associate with the value of a citizen science observation.
+We predicted that the number of days since the last sample would be positively associated with the value of a citizen science observation.
   \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site.
-We hypothesized that for the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
+We predicted that the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
   \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor.
-We hypothesized that this would be positively influence the value of a observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value observations.
+We predicted that the nearest-neighbor sampling interval would positively influence the value of an observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value.
   \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site.
-We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with lots of observations would receive value given the long-term data originating from them.
+We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with many observations would have additional value given the long-term data originating from them.
 \end{itemize}
@@
 Each sampling event is then assigned a cumulative measure of leverage --- i.e., marginal value.
-We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend detection of biodiversity.
-Finally, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given grid in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.
+We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend.
+To assess the robustness of our approach, we investigated our predictions at four different site scales, represented by grids of 5, 10, 25, and 50 km$^{2}$.
+
+Identifying the contribution of each parameter to marginal value, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given site in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.
 
 \section*{Results and Discussion}
-To assess the robustness of our approach, we investigated our predictions at four different grid sizes: 5, 10, 25, and 50 km$^{2}$ grids.
-We found weak evidence that visiting an unsampled grid was marginally more valuable than visiting an already sampled grid.
-There was no statistical significance for the 5 km (p=0.669), and 10 km (p=0.093) grid sizes, but there was for the 25 km (p=0.035) grid size.
-At the 50 km grid size, this test was not possible because all grids had been sampled.
-These results suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
-In other words, citizen scientists contributing to eBird, likely a large number who are qualified naturalists in their own right \cite{callaghan2018unnatural}, are already sufficiently sampling biodiversity in space.
+\subsection*{Tests of predictions}
+We found weak evidence that visiting an unsampled site was marginally more valuable than visiting an already sampled site.
+There was no statistical significance for the 5 km (p=0.669), and 10 km (p=0.093) site sizes, but there was for the 25 km (p=0.035) site size.
+At the 50 km site size, this test was not possible because all sites had been sampled.
+These results suggests that stratified sampling --- an approach which aims for equal sampling among sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
+In other words, citizen scientists are already sufficiently sampling biodiversity in space: they are appropriately identifying and sampling the 'hotspots' in space that should receive the most sampling attention.
 
@@
 
-For those sampled grids, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
+For those sampled sites, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
 2).
 In general, our parameters were positively associated with the marginal value of a sampling event.
-Of our parameters of interest, the number of unique days sampled (i.e., the `history' of a grid) had the strongest, positive, effect size, and this was robust to the grid size comparisons.
-The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km grid size.
-Distance to the nearest sampled grid and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters.
-Surprisingly, days since the last sample had less influence than the other parameters, albeit a positive association.
-The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, likely highlights the value of `revisiting' a site in order to extract the maximum amount of information.
+The number of unique days sampled --- likely represented from known 'hotspots' identified by citizen scientists --- had the strongest, positive, effect size, and this was robust to the site size comparisons.
+The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km site size.
+Distance to the nearest sampled site and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters.
+Surprisingly, the number of days since the last sample, while positively associated, had less influence than the other parameters.
+The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, highlights the value of `revisiting' a site (i.e., 'hotspot') in order to extract the maximum amount of information.
 
-Together, these results suggest that there are possible improvements in biodiversity sampling for trend detection.
-In short, the `history' of a grid is important for considering whether to sample that grid: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable.
-Further, a secondary goal should be to decrease (i.e., left-shift the distribution) of the median sampling interval of the grids.
-The effect of number of days sampled was strongest at the largest grid size, suggesting that given all grids are sampled with some regularity --- as they are with larger grid sizes --- the importance of the `history' of a site is inflated.
+Together, these results suggest that improvements can be made in biodiversity sampling for trend detection.
+The `history' of a site is particularly important while considering whether to sample that site: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable.
+A secondary goal should be to decrease (i.e., left-shift the distribution) the median sampling interval of the sites.
+The effect of number of days sampled was strongest at the largest site size, suggesting that given all sites are sampled with some regularity --- as they are with larger site sizes --- the importance of the `history' of a site is inflated.
 
-We found generally consistent results, based on the grid size chosen, highlighting the robustness of spatial scale in our approach.
+We found generally consistent results, based on the site size chosen, highlighting the robustness of spatial scale in our approach.
 It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}.
@@
 
+\subsection*{Applications of our predictions}
 Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}.
 This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
-We used our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
-3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}).
-This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and pedict the expected leverage for any given day.
-We envision a dynamic approach (Video S2) in the future of citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
-In this instance, leaderboards would move past numbers of species or submissions and complementary leaderboards could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
-Instead of participants preferrentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
+But leaderboards tend to focus on outputs, incentivizing finding rather than looking, leading to perverse outcomes related to the redundancies and gaps in citizen science data.
+We sought to develop an outcome-based incentive by using our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
+3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}).
+This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and predict the expected leverage for any given day.
+We envision a dynamic approach (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}) in future citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
+In this instance, leaderboards would move past numbers of species or submissions and could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
+Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
 For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science.
@@
 \includegraphics[width=.8\linewidth]{example_maps.png}
-\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued grid cells that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region.
-This is a dynamic approach, where the predictions are updated as new observations are submitted to the citizen science database, and we have done this for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video S2}).}
+\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued sites that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region.
+This prediction step is dynamic: the predictions are updated as new observations are submitted to the citizen science database.
+Prediction updating can be done at any frequency (see updates for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2})).
+Expected value maps will need to be updated fast in parts of the world where sampling rates are high, but this can be done at a slower rate where sampling is less frequent.}
 \label{fig3}
@@
 But these different outcomes can still be quantified in the same framework we introduce here.
-For example, an intended outcome of a spcecies distribution model would likely place greater value on observations from unsampled grids \cite{crawley2001scale} than for species trend detection.
-Our framework is robust, with the necessary piece of information being a statistical model based on an intended ecological/management outcome.
+For example, an intended outcome of a species distribution model would likely place greater value on observations from unsampled sites \cite{crawley2001scale} than for species trend detection.
+Our framework is robust, with the necessary information being a statistical model based on an intended ecological/management outcome.
 
@@
 And this is only one citizen science project.
-Clearly, the cumulative effort put-forth by citizen scientists is immense, and arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple centuries \cite{silvertown2009new} --- with an increasingly critical role \cite{mckinley2017citizen, pocock2018vision} in monitoring of biodiversity.
+Clearly, the cumulative effort put-forth by citizen scientists is immense, and arguably, citizen science will continue to shape the future of ecology and conservation --- as it has substantially for the past couple of centuries \cite{silvertown2009new} --- with an increasingly critical role in monitoring of biodiversity\cite{mckinley2017citizen, pocock2018vision}.
 But we need to look towards the future.
 Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future?
-We highlighted general rules which could help to guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation.
-Moreover, we demonstrate a framework which practicioners can implement to better optimize their sampling designs, specific to a citizen science project and an intended goal.
+We highlight general rules which could help guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation.
+Moreover, we demonstrate a framework which citizen science projects can implement to better optimize their sampling designs, which can be tailed to specific citizen science project goals.
 
-\matmethods{ We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$, where a grid represented a 'site'.
+\matmethods{We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$, where a grid represented a 'site'.
 We used the R statistical environment \cite{rcoreteam2018r} to carry out all analyses, relying heavily on the tidyverse \cite{wickham2017tidyverse}, ggplot2 \cite{wickham2016ggplot}, and sf \cite{pebesma2018sf} packages.
 
-In order to test our predictions, we used the eBird basic dataset (version $ebd-relDec-2018$; available at: https://ebird.org/data/download), subsetting the data between January 1$^{st}$, 2010 and December 31$^{st}$, 2018.
+In order to test our predictions, we used the eBird basic dataset (version $ebd-relDec-2018$; available at: \url{https://ebird.org/data/download}), subsetting the data between January 1$^{st}$, 2010 and December 31$^{st}$, 2018.
 eBird is a particularly successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
@@
 \subsection*{Statistical leverage} Statistical leverage measures the influence of a particular observation on the independent variable \cite{cook1977detection}.
-In other words, it is a measure of how much a given observation influences the results of a statistical outcome.
-In our instance, we had multiple predictor variables in our GLMs, and so we used dfBeta as a measure of statistical leverage for each observation.
-dfBeta measures the change to the observed parameters of a new regression equation, after omitting the \textit{i$^{th}$} observation from the dataset \cite{belsley1980regression}.
+In other words, it is a measure of how much a given observation affects the statistical model.
+In our instance, as is likely to be the case for all trend detection models, we had multiple predictor variables.
+Because for trend detection, we are interested in one specific model parameter--the coefficent of the probability of observation of that species with respect to time--dfBeta rather than Cook's D is appropriate \citep{bollinger1981book}.
+dfBeta measures the change to the one model parameter, after omitting the \textit{i$^{th}$} observation from the dataset \cite{belsley1980regression, bollinger1981book}.
 It follows the formula:
@@
 where X is the predictor variable matrix, \textit{r} the residual vector, \textit{i h} the \textit{i$^{th}$} diagonal member, and \textit{i x} the \textit{i$^{th}$} line of matrix X.
-The value of dfBeta proportionally decreases with increasing number of observations.
+The value of dfBeta tends to decline with an increase in the number of observations as the trend becomes well understood.
 
-In our case, each observation for a given species recieved a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the 'dfbetas' function from R \cite{rcoreteam2018r}.
+In our case, each observation for a given species received a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the 'dfbetas' function from R \cite{rcoreteam2018r}.
 The measure of statistical leverage, then, of a given checklist was the sum of the absolute value of the dfBeta measures for each species (i.e., the sum of all 235 dfBetas).
 This measure of statistical leverage was thus a measure of a checklist's influence in understanding cumulative species' trends throughout the Greater Sydney Region, and accordingly represented the marginal value of that particular checklist.
+Failing to observe a species produces a dfBeta which can be quite important in detecting a species decline.
 
@@
 For each individual grid, at each of the grid sizes, we dynamically calculated the following parameters: (1) whether a grid cell had ever been sampled, (2) the distance to the nearest sampled grid cell, (3) the median sampling interval of a grid cell, (4) the median sampling interval of the nearest sampled grid cell, (5) days since the last sample in a grid cell, and (6) the duration of sampling in a grid cell --- most recent sample minus the earliest sampled date, and (7) the number of unique sampling days within the grid cell.
+Note that  these parameters depend on the sampling in the days prior to that particular observation and do not consider the sampling in subsequent days.
 
 We then subsetted the leverage calculations (see above) for each of the days in 2018, given we know where people sampled, relative to the parameters for each of the grids on that day.
-We ran a linear regression, for each of the different grid sizes considered in the analysis, to investigate what parameters were of significant interest.
-Prior to modelling, duration was highly correlated with median sampling interval for the majority of the grid size analyses, and as such, was excluded from consdieration.
-Given the paramters' correlation varied among grid cell sizes, we needed to ensure a robust, and simple model.
+We ran a linear regression, for each of the different grid sizes considered in the analysis, to investigate which parameters could forecast checklist influence.
+Prior to modelling, duration was highly correlated with median sampling interval for the majority of the grid size analyses, and as such, was excluded from consideration.
+Given the parameters' correlation varied among grid cell sizes, we needed to ensure a robust, and simple model.
 All variables were log-transformed and then standardized prior to modelling, ensuring that the effect sizes of the given parameters were meaningful.
@@
 
-\subsection*{Data availability} All eBird data are freely available for download (https://ebird.org/data/download), but the necessary portion of the eBird basic dataset, along with spatial data, and code to reproduce our analyses are available at: https://github.com/coreytcallaghan/optimize_citizen_science_obs.
+\subsection*{Data availability} All eBird data are freely available for download (\url{https://ebird.org/data/download}), but the necessary portion of the eBird basic dataset, along with spatial data, and code to reproduce our analyses are available at: \url{https://github.com/coreytcallaghan/optimize_citizen_science_obs}.
 }

### 2019-04-25 R/plot_leverage.R
making an additional plot of 10 random grids' value through time

@@ -113,7 +113,33 @@ for (i in dates_2018) {
 }
 
 
+# plot the value through time for some example grids
+# this gives an additional figure and highlights the
+# 'dynamic' part of the paper a bit more
+sample_n_groups = function(tbl, size, replace = FALSE, weight=NULL) {
+  # regroup when done
+  grps = tbl %>% groups %>% unlist %>% as.character
+  # check length of groups non-zero
+  keep = tbl %>% summarise() %>% sample_n(size, replace, weight)
+  # keep only selected groups, regroup because joins change count.
+  # regrouping may be unnecessary but joins do something funky to grouping variable
+  tbl %>% semi_join(keep) %>% group_by_(grps) 
+}
 
+
+leverage_5 %>%
+  group_by(grid_id) %>%
+  sample_n_groups(10) %>%
+  ggplot(., aes(x=dynamic_date, y=Value, group=as.factor(grid_id), color=as.factor(grid_id)))+
+  geom_line()+
+  theme_bw()+
+  theme(axis.text=element_text(color="black"))+
+  theme(axis.title=element_text(color="black"))+
+  ylab("Marginal value")+
+  xlab("")+
+  labs(color="Grid ID")
+
+ggsave("Figures/example_grids_through_time.png", width=6.7, height=5, units="in")
 
 
 

### 2019-04-25 paper/main.tex
Adding additional figure to manuscript

@@
 
-\subsection*{Applications of our predictions}
-Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}.
-This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
-But leaderboards tend to focus on outputs, incentivizing finding rather than looking, leading to perverse outcomes related to the redundancies and gaps in citizen science data.
-We sought to develop an outcome-based incentive by using our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
-3, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}).
-This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and predict the expected leverage for any given day.
-We envision a dynamic approach (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}) in future citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
-In this instance, leaderboards would move past numbers of species or submissions and could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
-Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
-For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science.
-The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.
-
 \begin{figure}[!hb]
@@
 
+\subsection*{Applications of our predictions}
+Providing dynamic feedback to citizen science participants has proved successful for many citizen science projects \cite{rowley2019frogid, wiggins2011conservation, xue2016avicaching}.
+This feedback is generally in the form of leaderboards, presenting the number of submissions or number of unique species someone has contributed \cite{wood2011ebird}.
+But leaderboards tend to focus on outputs, incentivizing finding rather than looking, leading to perverse outcomes related to the redundancies and gaps in citizen science data.
+We sought to develop an outcome-based incentive by using our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day (e.g., Fig.
+3, Fig.
+4, \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}).
+This approach required us to look backward first, using a model with all observations for 2018, based on statistical leverage calculated from 2010-2018, in order to look forward and predict the expected leverage for any given day.
+We envision a dynamic approach (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2}) in future citizen science projects, which would ultimately guide participants to sites which \textit{should} be sampled on any given day --- or in a given week, month, or year.
+In this instance, leaderboards would move past numbers of species or submissions and could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
+Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
+For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science.
+The global pull of ecotourism \cite{sharpley2006ecotourism} is increasing exponentially, creating the potential for people to contribute to local biodiversity knowledge in areas that are traditionally undersampled, and with this framework, the collective effort of citizen scientists can be maximized.
+
 We focused our framework on a specific statistical outcome: trend detection.
@@
 Our framework is robust, with the necessary information being a statistical model based on an intended ecological/management outcome.
+
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=.8\linewidth]{example_grids_through_time.png}
+\caption{The changes in expected marginal value for 2018 for 10 randomly chosen grid cells.
+Some sites' expected value remains relatively constant through time (e.g., grid 371) while others are highly variable (e.g., grid 274), and others undergo distinct step-changes (e.g., grid 405).
+This is a dynamic approach, showing how the expected value of a grid changes based on our parameterized model results (Fig.
+3 \href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2})).}
+\label{fig4}
+\end{figure}
 

### 2019-05-01 paper/main.tex
updated paper with minor edits and tweaks. Submitted version

@@
 \significancestatement{The challenge of monitoring biodiversity during the ongoing extinction crisis is unlikely to be matched by sufficient funding to professional scientists.
-Citizen science may fill much of the shortfall, but currently, citizen science sampling is haphazard, and we can do a better job maximizing both the marginal value of each person's effort and the total information content of a given dataset.
+Citizen science may fill much of the shortfall, but currently, citizen science sampling is haphazard, and we can do a better job maximizing both the relative value of each person's effort and the total information content of a given dataset.
 We provide a dynamic framework to do this, showing where and when to sample biodiversity to maximize value per effort.
-We recommend that citizen science projects adapt this framework to their needs, with a focus on maximizing the marginal value of each citizen science contribution to inform biodiversity conservation goals.}
+We recommend that citizen science projects adapt this framework to their needs, with a focus on maximizing the value of each citizen science contribution to inform biodiversity conservation.}
 
@@
 \begin{abstract}
-We are currently in the midst of earth's sixth extinction event, and measuring biodiversity trends in space and time is essential for prioritizing limited resources for conservation and curbing biodiversity loss.
-At the same time, the scope of the necessary monitoring is massively overwhelming the funding for professional scientific monitoring.
-In response, scientists are increasingly using citizen science data to estimate population trends, with the aim of providing prioritizing conservation efforts.
-But citizen science data are 'noisy', with redundancies and gaps arising from unstructured human behaviors in space and time.
-Here, we ask whether the information in these data can be improved --- maximizing the information content extracted from citizen science datasets for trend estimation.
-We develop and execute a novel framework, based on statistical leverage, which assigns every citizen science observation a marginal value derived from the importance of an observation to our understanding of overall population trends for a given geographic region.
-We then make this framework predictive, estimating the expected marginal value of future biodiversity observations.
+We are currently in the midst of earth's sixth extinction event, and measuring biodiversity trends in space and time is essential for prioritizing limited resources for conservation.
+At the same time, the scope of the necessary biodiversity monitoring is overwhelming funding for professional scientific monitoring.
+In response, scientists are increasingly using citizen science data to monitor biodiversity.
+But citizen science data are `noisy', with redundancies and gaps arising from unstructured human behaviors in space and time.
+We ask whether the information content of these data can be maximized for the express purpose of trend estimation.
+We develop and execute a novel framework which assigns every citizen science sampling event a relative value, derived from the importance of an observation to our understanding of overall population trends.
+We then make this framework predictive, estimating the expected value of future biodiversity observations.
 We find that past observations are useful in forecasting where high value observations will occur in the future.
-Interestingly, we find high marginal value in both `hotspots` which are frequently sampled locations and `coldspots` areas far from recent sampling, suggesting that an optimal sampling regime balances `hotspot` sampling with a spread across the landscape.
-Our framework is applicable at any spatial scale, easily adapted within a national park, entire country, or global.
-Because expected value is dependent on the behavior of all other citizen scientists, our model produces dynamic maps of expected marginal value through time.
-Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected marginal value, collectively maximizing our understanding of biodiversity in space and time, and our ability to conserve it.
+Interestingly, we find high value in both `hotspots' which are frequently sampled locations and `coldspots' which are areas far from recent sampling, suggesting that an optimal sampling regime balances `hotspot' sampling with a spread across the landscape.
+Our framework is applicable at any spatial scale, easily adapted within a national park, entire country, or globally.
+Ultimately, we envision an approach where citizen scientists are guided and incentivized to sample sites with the highest expected value, collectively maximizing our understanding of biodiversity in space and time, and hence our ability to conserve it.
 \end{abstract}
@@
 
-Citizen science --- the cooperation between scientific experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}.
-Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects which require little to no training required; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K.
+Citizen science --- the cooperation between scientific experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen} and is rapidly expanding \cite{newman2012future, pocock2017diversity}.
+Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., opportunistic or incidental projects which require little to no training; iNaturalist) to structured (e.g., projects with specific objectives, rigorous protocols, and survey design; U.K.
 Butterfly Monitoring Scheme).
@@
 Generalizeable among citizen science projects, however, are various redundancies and gaps (i.e., spatial and temporal biases) \cite{boakes2010distorted, bird2014statistical}.
-Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and visiting known 'hotspots' for biodiversity \cite{geldmann2016determines} are all examples of how unstructured human behavior leads to redundancies and gaps in citizen science data.
+Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and visiting known `hotspots' for biodiversity \cite{geldmann2016determines} are all examples of how unstructured human behavior leads to redundancies and gaps in citizen science data.
 These biases are not restricted to citizen science projects.
-Indeed, our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted} and many sampling methods have been proposed to optimize biodiversity sampling by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
+Indeed, our historical understanding of biodiversity in space and time is also biased; evidenced by natural history collections \cite{pyke2010biological, boakes2010distorted}, and many sampling methods have been proposed to optimize biodiversity sampling by professional scientists \cite{etienne2005new, moreno2000assessing, colwell1994estimating, longino1997biodiversity, ferrarini2012biodiversity}, frequently dependent on spatial scale \cite{chase2013scale}.
 While structured citizen science projects often adapt some aspects of optimal sampling in their methods (e.g., stratified sampling), little attention has been given to optimal sampling in unstructured citizen science projects \cite{harrison2014assessing}.
@@
 But unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.
-The robustness of these trend estimates is critical, and the goal should be to continuously decrease the standard error surrounding these estimates (e.g., Fig.
+The robustness of these trend estimates is critical, and the goal should be to continuously decrease the uncertainty surrounding these estimates (e.g., Fig.
 1).
-Unsurprisingly, the standard error is generally proportional to the number of observations, as well as appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video 1}).
+Unsurprisingly, uncertainty is generally proportional to the number of observations, as well as appropriate sampling, through time (e.g.,\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/Noisy_miner_gif.gif}{Video 1}).
 
@@
 \includegraphics[width=.8\linewidth]{glm_examples_edited.png}
-\caption{The ultimate goal in understanding population trends is to minimize the standard error for a population trend model, providing more robust measures of population trends.
-Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right), in the Greater Sydney Region, Australia.}
+\caption{The ultimate goal in understanding population trends is to minimize the uncertainty for a population trend model, providing more robust measures of population trends.
+Shown here are four example population trend models, based on eBird data between 2010-2018, for Noisy Miner (top left), Hardhead (top right), Masked Lapwing (bottom left), and Crested Pigeon (bottom right), in the Greater Sydney Region, Australia.
+Each model incorporates \textasciitilde 26,000 biodiversity sampling events.}
 \label{fig1}
@@
 The number of citizen science projects which are focused on ecological and environmental monitoring is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}.
-But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.
-As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using 'big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data.
-And even with 'big data', this is not always possible.
+But a major obstacle in the future use of citizen science data remains understanding how to best extract information from `noisy' citizen science datasets \cite{parrish2018exposing}.
+As mentioned, this noise from citizen science \cite{bird2014statistical} can sometimes be alleviated using `big-data' statistical approaches \cite{kelling2015taking}, but this is most applicable for data originating from large, successful citizen science projects --- with lots of data.
+And even with big-data, this is not always possible.
 But what about projects that are just beginning?
@@
 Here, we investigate these questions with a specific objective: assess how spatial and temporal sampling influences trend detection of biodiversity.
-Our approach is dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation in both time and space.
-First, we summarize the parameters which we hypothesized would influence the marginal value of a citizen science observation.
+Our approach is dynamic: we are interested in the parameters that influence the value of a given citizen science sampling event in both time and space.
+``Value'' is assigned by using statistical leverage: a measure of the influence a given observation has on the population trend model of a species.
+Each sampling event --- unique in time and space --- is then assigned a cumulative measure of leverage (i.e., relative value) by summing the influence that the sampling event has on all 235 species' trends.
+This can then be thought of as the cumulative value of a sampling event to inform our total knowledge of species' trends, across many species.
+First, we summarize the parameters which we hypothesized would influence the relative value of a citizen science sampling event.
 
@@
 We fit presence/absence population trend models to 235 species throughout this region, based on data from 2010 to 2018 (Fig.
-1), and calculate statistical leverage for every citizen science sampling event, for each species.
-The leverage is a measure of how 'valuable' a given observation is to the population trend model for each species.
-Each sampling event is then assigned a cumulative measure of leverage --- i.e., marginal value.
-We then test our predictions above using linear models, assessing the influence each of the parameters has on the trend.
+1), and calculate statistical leverage for every citizen science sampling event, taken as the cumulative influence a sampling event has across all species, accounting for presences and absences derived from a sampling event.
+We then test our predictions above using linear models, assessing the influence each of the parameters has on uncertainty surrounding population trends.
 To assess the robustness of our approach, we investigated our predictions at four different site scales, represented by grids of 5, 10, 25, and 50 km$^{2}$.
 
-Identifying the contribution of each parameter to marginal value, we use the parameterized model results to predict the expected leverage (i.e., marginal value) for a given site in space and time, producing a dynamic map of expected marginal value of citizen science sampling for every day in 2018.
+After identifying the contribution of each parameter to relative value, we use the parameterized model results to predict the expected leverage (i.e., expected value) for a given site in space and time, producing a dynamic map of expected value of citizen science sampling for every day in 2018.
 
@@
 These results suggests that stratified sampling --- an approach which aims for equal sampling among sites --- \cite{baillie2008toward, longino1997biodiversity} is not the most applicable approach for detecting trends using citizen science data.
-In other words, citizen scientists are already sufficiently sampling biodiversity in space: they are appropriately identifying and sampling the 'hotspots' in space that should receive the most sampling attention.
+In other words, citizen scientists are already sufficiently sampling biodiversity in space: they are appropriately identifying and sampling the `hotspots' in space that should receive the most sampling attention.
 
@@
 \includegraphics[width=.8\linewidth]{param_estimates.png}
-\caption{Parameter estimates for standardized linear models for each of the four grid sizes (5, 10, 25, and 50).
-Variables were log-transformed and then standardized, allowing for the direct comparison among effect sizes.}
+\caption{Parameter estimates for standardized linear models for each of the four grid sizes investigated (5, 10, 25, and 50 km$^{2}$).
+Variables were log-transformed and then standardized, allowing for direct comparison among effect sizes.}
 \label{fig2}
@@
 
-For those sampled sites, however, we found significant relationships between a number of our predicted parameters and the marginal value of a sampling event (Fig.
+For those sampled sites, however, we found significant relationships between a number of our predicted parameters and the relative value of a sampling event (Fig.
 2).
-In general, our parameters were positively associated with the marginal value of a sampling event.
-The number of unique days sampled --- likely represented from known 'hotspots' identified by citizen scientists --- had the strongest, positive, effect size, and this was robust to the site size comparisons.
-The median sampling interval was also strongly associated with high marginal value samples, with an exception at the 25 km site size.
-Distance to the nearest sampled site and the nearest neighbor sampling interval influenced the marginal value of a sampling event less than the other parameters.
+In general, our parameters were positively associated with the relative value of a sampling event.
+The number of unique days sampled --- likely represented from known `hotspots' identified by citizen scientists --- had the strongest, positive, effect size, and this was robust to the site-size comparisons.
+The median sampling interval was also strongly associated with high value samples, with an exception at the 25 km site size.
+Distance to the nearest sampled site and the nearest neighbor sampling interval influenced the value of a sampling event less than the other parameters.
 Surprisingly, the number of days since the last sample, while positively associated, had less influence than the other parameters.
-The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, highlights the value of `revisiting' a site (i.e., 'hotspot') in order to extract the maximum amount of information.
+The fact that days since last sample had a lower effect size than both the median sampling interval and the number of unique days sampled, highlights the value of "revisiting" a site (i.e., "hotspot") in order to extract the maximum amount of information.
 
 Together, these results suggest that improvements can be made in biodiversity sampling for trend detection.
-The `history' of a site is particularly important while considering whether to sample that site: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are marginally more valuable.
+The "history" of a site is particularly important while considering whether to sample that site: the number of unique days sampled was the strongest predictor, suggesting that observations from sites with a long time-series are relatively more valuable.
 A secondary goal should be to decrease (i.e., left-shift the distribution) the median sampling interval of the sites.
-The effect of number of days sampled was strongest at the largest site size, suggesting that given all sites are sampled with some regularity --- as they are with larger site sizes --- the importance of the `history' of a site is inflated.
+The effect of number of days sampled was strongest at the largest site size, suggesting that given all sites are sampled with some regularity --- as they are with larger site sizes --- the importance of the "history" of a site is inflated.
 
-We found generally consistent results, based on the site size chosen, highlighting the robustness of spatial scale in our approach.
+We found generally consistent results, based on the size of the site chosen, highlighting the robustness of spatial scale in our approach.
 It is critical to track biodiversity trends at multiple spatial scales \cite{soberon2007assessing}, as biodiversity estimates sometimes change dependent on the spatial scale \cite{chase2013scale}.
@@
 \includegraphics[width=.8\linewidth]{example_maps.png}
-\caption{A map of predicted expected marginal value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued sites that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region.
-This prediction step is dynamic: the predictions are updated as new observations are submitted to the citizen science database.
+\caption{A map of predicted expected value for six different days in 2018, throughout the Greater Sydney Region, showing the highest valued sites that would optimize the collective knowledge on biodiversity trends throughout the Greater Sydney Region.
+This prediction step is dynamic: predictions are updated as new observations are submitted to the citizen science database.
 Prediction updating can be done at any frequency (see updates for every day in 2018 (\href{https://github.com/coreytcallaghan/optimize_citizen_science_obs/blob/master/Figures/dynamic_map.gif}{Video 2})).
-Expected value maps will need to be updated fast in parts of the world where sampling rates are high, but this can be done at a slower rate where sampling is less frequent.}
+Expected value maps will need to be updated fast in parts of the world where sampling rates are high, but this can be done at a slower rate where sampling is less frequent,and this will also vary among projects.}
 \label{fig3}
@@
 In this instance, leaderboards would move past numbers of species or submissions and could be derived based on a participant's cumulative value to the citizen science dataset, for biodiversity outcomes.
-Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected marginal value.
+Instead of participants preferentially chasing specific species, this approach would guide participants to the sites with the highest expected value.
 For example, we imagine visitor centers across the world at national parks or urban greenspaces providing their visitors on any given day a localized map showing which trail someone should visit if they are interested in contributing the greatest value to that park's biodiversity knowledge, through citizen science.
@@
 \includegraphics[width=.8\linewidth]{example_grids_through_time.png}
-\caption{The changes in expected marginal value for 2018 for 10 randomly chosen grid cells.
-Some sites' expected value remains relatively constant through time (e.g., grid 371) while others are highly variable (e.g., grid 274), and others undergo distinct step-changes (e.g., grid 405).
+\caption{The changes in expected value for 2018 for 10 randomly chosen grid cells at the 5 km$^{2}$ grid size.
+Some sites' expected value remains relatively constant through time (e.g., grid 371) while others are highly variable (e.g., grid 274), and others undergo distinct step-changes (e.g., grid 405) corresponding to whether that grid was recently sampled or not.
 This is a dynamic approach, showing how the expected value of a grid changes based on our parameterized model results (Fig.
@@
 Are there mechanisms we can put in place now which will increase our collective knowledge gleaned from citizen science datasets for biodiversity in the future?
-We highlight general rules which could help guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the marginal value of a citizen science observation.
-Moreover, we demonstrate a framework which citizen science projects can implement to better optimize their sampling designs, which can be tailed to specific citizen science project goals.
+We highlight general rules which could help guide citizen science participants to better sampling in space and time: the number of unique days sampled and the largest median sampling intervals both positively correlate with the relative value of a citizen science observation.
+Moreover, we demonstrate a framework which citizen science projects can implement to better optimize their sampling designs, which can be tailored to specific citizen science project goals.
 
-\matmethods{We tested our predictions throughout the Greater Sydney Region, delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$, where a grid represented a 'site'.
+\matmethods{We tested our predictions throughout the Greater Sydney Region (\textasciitilde 12,400 km$^{2}$), delineating grids across the region of varying size: 5, 10, 25, and 50 km$^{2}$, where a grid represented a `site'.
 We used the R statistical environment \cite{rcoreteam2018r} to carry out all analyses, relying heavily on the tidyverse \cite{wickham2017tidyverse}, ggplot2 \cite{wickham2016ggplot}, and sf \cite{pebesma2018sf} packages.
@@
 eBird is a particularly successful citizen science project with > 600 million observations contributed by > 400 thousand participants, globally \cite{sullivan2009ebird, sullivan2014ebird, sullivan2017using}.
-eBird relies on volunteer birdwatchers who collect data in the form of 'checklists' --- a list of all species identified (audibly or visually) for given spatiotemporal coordinates.
+eBird relies on volunteer birdwatchers who collect data in the form of `checklists' --- a list of all species identified (audibly or visually) for given spatiotemporal coordinates.
 eBird relies on an extensive network of regional reviewers who are local experts of the avifauna \cite{gilfedder2019brokering} to ensure data quality \cite{sullivan2009ebird}.
@@
 
-For any species with > 50 observations (N=235), we fit a generalized linear model using the 'glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}.
+For any species with > 50 observations (N=235), we fit a generalized linear model using the `glm' function in R, based on presence/absence \cite{walker2017using, horns2018using}.
 The models consisted of a continuous term for day, beginning January 1$^{st}$, 2010, and a categorical term for county, providing a spatial component to the models (e.g., Fig.
@@
 Each of the models were fit with a binomial family distribution.
-A total of 25,995 observations (i.e., eBird checklists) were used to fit each of the models.
+A total of 25,995 sampling events (i.e., eBird checklists) were used to fit each of the models.
 
@@
 
-In our case, each observation for a given species received a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the 'dfbetas' function from R \cite{rcoreteam2018r}.
+In our case, each observation for a given species received a dfBeta value (i.e., each species received 25,995 measures of dfBeta), using the `dfBetas' function from R \cite{rcoreteam2018r}.
 The measure of statistical leverage, then, of a given checklist was the sum of the absolute value of the dfBeta measures for each species (i.e., the sum of all 235 dfBetas).
@@
 
-After our model was fit, we used the 'augment' function from the broom package \cite{robinson2018broom} to predict the expected leverage for every grid cell in the Greater Sydney Region, for every day.
+After our model was fit, we used the `augment' function from the broom package \cite{robinson2018broom} to predict the expected leverage for every grid cell in the Greater Sydney Region, for every day.
 For grid cells which were unsampled, we assigned them the mean of the sampled grid cells, based on our lack of evidence that unsampled cells were significantly more valuable than sampled cells.
@@
 
-\acknow{We thank the countless citizen scientists who are contributing data that is continuously increasing our collective knowledge of biodiversity.}
+\acknow{We thank the countless citizen scientists who are contributing data that is continuously increasing our collective knowledge of biodiversity.
+We also thank the eBird team and the Cornell Lab of Ornithology who are working to further bird data collected by citizen scientists and making that data publicly available.}
 
